### 2024-10-29 paper/paper.tex
add latex-template

\documentclass[sigplan,11pt,nonacm]{acmart}
\settopmatter{printfolios}

\usepackage{booktabs} % For formal tables
\usepackage{subcaption}
\usepackage{tikz}
\usepackage{pgfplots}
\usepackage{pgfplotstable}
\usepackage{hyphenat}
\usepackage{todonotes}
\usepackage[babel]{csquotes}

\begin{document}
\title{History of Beach Computers}
\author{Best Student}
\author{Better Student}
\affiliation{%
  \institution{Technical University of Munich}
}
\email{{best,better}@example.com}

\begin{abstract}
Your Abstract.
\end{abstract}

% If you like to...
%\keywords{Beach, Computers, History}

\maketitle

\section{Introduction}
\label{sec:introduction}

Beaches are a very natural part of countries and are especially useful for vacation. However, even when on travel, programmers obviously do not want to live without computers. This led to the development of \emph{beach computers}. This type of computers was introduced in 1970 by Some Inventor~\cite{intel2019man} % <-- some totally random reference
to satisfy needs of programmers, but also to boost increase their vacation productivity. \textit{To be continued.}

\section{Summary \& Outlook}
\label{sec:summary}

Oh, and here goes the summary. Future work is expected in water computers so that programmers can also be productive while swimming.

\bibliographystyle{ACM-Reference-Format}
\bibliography{paper} % read paper.bib file

\end{document}

### 2024-11-06 paper/paper.tex
rough paper structure #1

@@
 \begin{document}
-\title{History of Beach Computers}
-\author{Best Student}
-\author{Better Student}
+\title{Utilizing Parallel Workers: \\LLVM's Vectorization Plan}
+\author{Jonas Fritsch}
 \affiliation{%
@@
 }
-\email{{best,better}@example.com}
+\email{jonas.fritsch@tum.de}
 
@@
 
-% If you like to...
-%\keywords{Beach, Computers, History}
+\maketitle
 
-\maketitle
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 
-Beaches are a very natural part of countries and are especially useful for vacation.
-However, even when on travel, programmers obviously do not want to live without computers.
-This led to the development of \emph{beach computers}.
-This type of computers was introduced in 1970 by Some Inventor~\cite{intel2019man} % <-- some totally random reference
-to satisfy needs of programmers, but also to boost increase their vacation productivity. \textit{To be continued.}
+\begin{enumerate}
+  \item 
+\end{enumerate}
 
-\section{Summary \& Outlook}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\section{Background}
+\label{sec:background}
+\begin{enumerate}
+  \item (SIMD explanation necessary?)
+  \item Vectorization explanation (with Code-Examples?)
+  \begin{enumerate}
+    \item Inner-Loop
+    \item Outer-Loop
+    \item Superword Level Parallelism~\cite{10.1145/349299.349320}~(maybe already \cite{10.1145/3519939.3523701})
+  \end{enumerate}
+  \item Vectorization Drawbacks (reuse code-examples from above)
+  \begin{itemize}
+    \item possible performance loss
+    \item larger code size
+    \item (potential security vulnerablities~\cite{2023arXiv230201131K})
+  \end{itemize}
+  \item LLVM short introduction
+\end{enumerate}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\section{LLVM's Vectorization Plan}
+\label{sec:vplan}
+\begin{enumerate}
+  \item Cost-Model (related work?~\cite{POHL2020102106})
+\end{enumerate}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\section{Related Work}
+\label{sec:relatedwork}
+\begin{enumerate}
+  \item Auto-Vectorization in GCC~\cite{autovecoptgcc, vecoptgcc}
+  \item Comparison with GCC~\cite{gccllvmveccomp,10.1155/2021/3264624}
+  \item (LLM-based Vectorization~\cite{taneja2024llmvectorizerllmbasedverifiedloop}, also if included maybe better fitting in "Future Work")
+\end{enumerate}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\section{Summary and Future Work}
 \label{sec:summary}
+\begin{enumerate}
+  \item VPlan Summary and Future Roadmap~\cite{llvmvplan}
+  \item Comparison/Outlook auto-vectorization vs. manual vectorization~\cite{10.1155/2021/3264624,9802745}
+\end{enumerate}
 
-Oh, and here goes the summary. Future work is expected in water computers so that programmers can also be productive while swimming.
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 

### 2024-11-11 paper/paper.tex
rough paper structure #2

@@
 \begin{abstract}
-Your Abstract.
+Lorem ipsum.
 \end{abstract}
@@
 \begin{enumerate}
-  \item 
+  \item Modern CPUs (SIMD-Registers, SIMD ISAs)
+  \item SIMD performance gains
+  \item Auto-Vectorization to help programmers better utilize SIMD functionality out of the box
+  \item LLVM as popular, platform-agnostic compiler framework
+  \item auto-vectorization complexity $\rightarrow$ wish for well designed, 
+  interoperable system (VPlan) $\rightarrow$ ongoing 8 year long effort started by Intel
 \end{enumerate}
@@
 \begin{enumerate}
-  \item (SIMD explanation necessary?)
-  \item Vectorization explanation (with Code-Examples?)
-  \begin{enumerate}
-    \item Inner-Loop
-    \item Outer-Loop
-    \item Superword Level Parallelism~\cite{10.1145/349299.349320}~(maybe already \cite{10.1145/3519939.3523701})
-  \end{enumerate}
-  \item Vectorization Drawbacks (reuse code-examples from above)
+  \item Vectorization explanation (with Code\hyp{}Examples)
   \begin{itemize}
-    \item possible performance loss
-    \item larger code size
+    \item Loop-Level Vectorization (Inner / Outer-Loops~\cite{llvmextloopvec,intelouterloop,10.5555/2337159.2337210,llvmouterloop,10.1145/1454115.1454119})
+    \item Function Vectorization~\cite{llvmextloopvec}
+    \item Superword-Level Parallelism\\Vectorization~\cite{10.1145/349299.349320}~(maybe already \cite{10.1145/3519939.3523701})
+  \end{itemize}
+  \item Vectorization Drawbacks~\cite{llvmveccontrolflow,intelvecessen} (reuse code\hyp{}examples from above)
+  \begin{itemize}
+    \item not always legal / applicable, e.g.: backwards data-dependencies, pointer aliasing, 
+    platform backwards-compatability / support, control flow complexity (function calls, register pressure), FP inaccuracies (\texttt{-ffast-math})
+    \item possible performance loss, e.g.: costly conversions (horizontal aggregations, integer division) resulting in slower overall code
+    on average
+    \item larger code size, e.g.: need for scalar loop-tail / epilogue
     \item (potential security vulnerablities~\cite{2023arXiv230201131K})
   \end{itemize}
-  \item LLVM short introduction
+  \item OpenMP (very short)
+  \item Auto-Vectorization in Compilers (short)
+  \item LLVM (very brief introduction)~\cite{10.5555/977395.977673}
+  \item LLVM 'recent' transition from Loop Vectorizer + SLP Vectorizer to 
+  Vectorization Plan (ongoing 8 year long refactoring and improvment 
+  effort started by Intel)
 \end{enumerate}
@@
 \label{sec:vplan}
+  Overview~\cite{llvmvec,llvmvplan,llvmextloopvec,llvmintrvplan,llvmvplanstate,llvmvplanrv,llvmouterloop,llvmvplanupdate}
 \begin{enumerate}
-  \item Cost-Model (related work?~\cite{POHL2020102106})
+  \item Phase 1: Legality
+  \item Phase 2: Planning (Vectorization Plans)
+  \begin{enumerate}
+    \item Building initial VPlan based on Phase 1 Constraints (Generate one or more up-to-date Code Examples for VPlan similar to~\cite{llvmintrvplan,llvmvplanstate,llvmouterloop})
+    $\rightarrow$ Explain different components (e.g.: \texttt{VPBasicBlock}, \texttt{VPRegionBlock}, etc.) and
+    correlation between generated VPlan structure, original scalar code. Show some general optimizations.
+    \item Modelling Cost (related work? already adopted?~\cite{POHL2020102106})
+    \item Optimizing/Transforming VPlans
+  \end{enumerate}
+  \item Phase 3: Execution - Materialize best VPlan
 \end{enumerate}
@@
 \begin{enumerate}
+  \item Other auto-vectorization algorithms / possible improvements~\cite{10.5555/3615924.3615932,4336219,llvmveccontrolflow,saarlandrv,10.1145/3303117.3306172}
+  \\Some of those should already be (partially) implemented in LLVM
   \item Auto-Vectorization in GCC~\cite{autovecoptgcc, vecoptgcc}
   \item Comparison with GCC~\cite{gccllvmveccomp,10.1155/2021/3264624}
-  \item (LLM-based Vectorization~\cite{taneja2024llmvectorizerllmbasedverifiedloop}, also if included maybe better fitting in "Future Work")
+  \item (Polyhedral compilation techniques~\cite{llvmpolyhedral})
+  \item (LLM-based Vectorization~\cite{taneja2024llmvectorizerllmbasedverifiedloop})
 \end{enumerate}
@@
 \begin{enumerate}
-  \item VPlan Summary and Future Roadmap~\cite{llvmvplan}
-  \item Comparison/Outlook auto-vectorization vs. manual vectorization~\cite{10.1155/2021/3264624,9802745}
+  \item VPlan Summary~\cite{llvmvplan,llvmvplanupdate}
+  \item VPlan Future Roadmap~\cite{llvmvplan,llvmvplanupdate} (e.g.: combining outer-/inner-loop paths, full function vectorization, etc.)
+  \item (Comparison/Outlook auto-vectorization and explicit vectorization~\cite{10.1155/2021/3264624, 9802745, DBLP:conf/vldb/BensonER23})
+  (\cite{chinese180}~only in chinese?)
 \end{enumerate}

### 2024-11-11 paper/paper.tex
rough paper structure #3

@@
   interoperable system (VPlan) $\rightarrow$ ongoing 8 year long effort started by Intel
+  \item rough paper outline
 \end{enumerate}
@@
     correlation between generated VPlan structure, original scalar code. Show some general optimizations.
-    \item Modelling Cost (related work? already adopted?~\cite{POHL2020102106})
-    \item Optimizing/Transforming VPlans
+    \item Modelling Cost. Not yet part of VPlan. (related work? already adopted?~\cite{POHL2020102106})
+    \item Optimizing/Transforming VPlans~\cite{llvmvplanupdate}
   \end{enumerate}
@@
   \item (Comparison/Outlook auto-vectorization and explicit vectorization~\cite{10.1155/2021/3264624, 9802745, DBLP:conf/vldb/BensonER23})
-  (\cite{chinese180}~only in chinese?)
 \end{enumerate}

### 2024-11-16 code/simple_loop.cpp
add code

@@ -1,0 +1,34 @@
+#define ENABLE_VECTORIZATION 1
+
+#if ENABLE_VECTORIZATION
+    #define VECTORIZABLE_LOOP \
+    _Pragma("clang loop vectorize(enable) interleave(enable) vectorize_predicate(enable)")
+#else
+    #define VECTORIZABLE_LOOP \
+    _Pragma("clang loop vectorize(disable) interleave(disable)")
+#endif
+
+void simple_loop_restrict(const int N, float *__restrict x, const float *__restrict y, const float *__restrict z) {
+    VECTORIZABLE_LOOP
+    for (int i = 0; i < N; i++) {
+        x[i] = y[i] + z[i];
+    }
+}
+
+//void simple_loop(const int N, float* x, const float* y, const float* z) {
+//    VECTORIZABLE_LOOP
+//    for (int i = 0; i < N; i++) {
+//        x[i] = y[i] + z[i];
+//    }
+//}
+
+// void outer_loop(const int N, const int M, const bool cond, float** a, float** b, float** c) {
+//     VECTORIZABLE_LOOP
+//     for (int i = 0; i < N; i++) {
+//         if (cond) {
+//             for (int j = 0; j < M; j++) {
+//                 a[j][i] = b[j][i] + c[j][i];
+//             }
+//         }
+//     }
+// }

### 2024-11-18 paper/outline.tex
add outline as seperate document

--- 2024-11-11 paper/paper.tex
@@
   \item SIMD performance gains
-  \item Auto-Vectorization to help programmers better utilize SIMD functionality out of the box
-  \item LLVM as popular, platform-agnostic compiler framework
+  \item Auto-Vectorization to help programmers utilize SIMD functionality out of the box 
+  (explicit vectorization as time consuming especially when supporting multiple architectures)
+  \item LLVM as popular, platform-agnostic compiler framework~\cite{10.5555/977395.977673}
   \item auto-vectorization complexity $\rightarrow$ wish for well designed, 
-  interoperable system (VPlan) $\rightarrow$ ongoing 8 year long effort started by Intel
+  interoperable system (VPlan) $\rightarrow$ transition from Loop Vectorizer + SLP Vectorizer to 
+  Vectorization Plan $\rightarrow$ ongoing 8 year long effort started by Intel
   \item rough paper outline
@@
   \end{itemize}
-  \item Vectorization Drawbacks~\cite{llvmveccontrolflow,intelvecessen} (reuse code\hyp{}examples from above)
+  \item Vectorization Constraints~\cite{llvmveccontrolflow,intelvecessen} (reuse code\hyp{}examples from above)
   \begin{itemize}
@@
   \end{itemize}
-  \item OpenMP (very short)
-  \item Auto-Vectorization in Compilers (short)
-  \item LLVM (very brief introduction)~\cite{10.5555/977395.977673}
-  \item LLVM 'recent' transition from Loop Vectorizer + SLP Vectorizer to 
-  Vectorization Plan (ongoing 8 year long refactoring and improvment 
-  effort started by Intel)
 \end{enumerate}

### 2024-11-18 paper/paper.tex
add outline as seperate document

--- 2024-11-06 paper/paper.tex
@@
 \begin{abstract}
-Your Abstract.
+Lorem ipsum
 \end{abstract}
@@
 \label{sec:introduction}
-
-\begin{enumerate}
-  \item 
-\end{enumerate}
+Lorem ipsum
 
@@
 \label{sec:background}
-\begin{enumerate}
-  \item (SIMD explanation necessary?)
-  \item Vectorization explanation (with Code-Examples?)
-  \begin{enumerate}
-    \item Inner-Loop
-    \item Outer-Loop
-    \item Superword Level Parallelism~\cite{10.1145/349299.349320}~(maybe already \cite{10.1145/3519939.3523701})
-  \end{enumerate}
-  \item Vectorization Drawbacks (reuse code-examples from above)
-  \begin{itemize}
-    \item possible performance loss
-    \item larger code size
-    \item (potential security vulnerablities~\cite{2023arXiv230201131K})
-  \end{itemize}
-  \item LLVM short introduction
-\end{enumerate}
+Lorem ipsum
 
@@
 \label{sec:vplan}
-\begin{enumerate}
-  \item Cost-Model (related work?~\cite{POHL2020102106})
-\end{enumerate}
+Lorem ipsum
 
@@
 \label{sec:relatedwork}
-\begin{enumerate}
-  \item Auto-Vectorization in GCC~\cite{autovecoptgcc, vecoptgcc}
-  \item Comparison with GCC~\cite{gccllvmveccomp,10.1155/2021/3264624}
-  \item (LLM-based Vectorization~\cite{taneja2024llmvectorizerllmbasedverifiedloop}, also if included maybe better fitting in "Future Work")
-\end{enumerate}
+Lorem ipsum
 
@@
 \label{sec:summary}
-\begin{enumerate}
-  \item VPlan Summary and Future Roadmap~\cite{llvmvplan}
-  \item Comparison/Outlook auto-vectorization vs. manual vectorization~\cite{10.1155/2021/3264624,9802745}
-\end{enumerate}
+Lorem ipsum
 

### 2024-11-19 paper/paper.tex
paper

@@
 \label{sec:introduction}
-Lorem ipsum
+Modern CPUs are often equipped with multiple different vector registers. These registers can nowadays 
+be as wide as 512 bit, allowing for the processing of multiple data streams at once (SIMD). By 
+batching multiple values together in one register, different ISAs like Intel AVX or ARM SVE allow 
+the execution of the same instruction for all values in a vector register at once. Utilizing this 
+can often lead to significant performance improvements over the scalar equivalent.
+
+However, as manual code vectorization can quickly become very time consuming especially when
+supporting different CPU architectures, modern compilers aim to automatically transform scalar code
+to use vectorization when applicable.
+
+As one of the most widely used compilation frameworks LLVM~\cite{10.5555/977395.977673} had 
+implemented and refined its auto-vectorization over many years. It provides two different 
+Vectorizers, one for innermost loops (LoopVectorize) and one for super-word parallelism 
+(SLPVectorize)~\cite{llvmvec}.
+
+This system however had quite a few limitations, as the loop vectorizer could only handle
+innermost loops and neither outer loops, complex control flow, or non-inlined functions. 
+Additionally, while multiple different vectorizatons for the same scalar code 
+would be possible, the current vectorizers working directly on the LLVM IR had no capability 
+of modelling and comparing the costs of such different vectorization approaches.
+
+With these limitations in mind Intel started an ongoing refactorization effort to migrate LLVM's
+auto-vectorization pipeline to use a more abstract Vectorization Plan 
+(VPlan)~\cite{llvmextloopvec,llvmvplan}. The final goal would be to unite LLVM's auto-vectorization
+in a single flexible system capable of optimizing SLPs, inner and nested loops with complex 
+control flows. The auto-vectorization pass would create, compare, and transform multiple different 
+VPlans each modelling a different vectorization approach and finally materialize the best one into
+the LLVM IR.
+
+In its current state LLVM's VPlan system is being integrated into the existing Loop Vectorizer and
+already models most of the inner loop vectorization and transformations. Vectorization for outer 
+loops is also in development and can be enabled by setting the \texttt{-enable-vplan-native-path} 
+flag ~\cite{llvmouterloop}. In the future the plan will be to merge both of these loop vectorization
+paths into one.
 

### 2024-11-19 paper/paper.tex
paper

--- 2024-11-18 paper/paper.tex
@@
 \label{sec:introduction}
-Lorem ipsum
+Modern CPUs are often equipped with multiple different vector registers. These registers are nowadays 
+as wide as 512 bits, allowing for the processing of multiple data streams at once (SIMD). By 
+batching multiple values together in one register, different ISAs like Intel AVX or ARM SVE allow 
+the execution of the same instruction for all values in a vector register simultaneously. 
+Utilizing this leads to significant performance improvements over the scalar equivalent most of 
+the time.
+
+However, as manual code vectorization can quickly become very time-consuming, especially when
+supporting different CPU architectures, modern compilers aim to automatically transform scalar code
+to use vectorization when applicable.
+
+As one of the most widely used compilation frameworks, LLVM~\cite{10.5555/977395.977673} has 
+implemented and refined its auto-vectorization over many years. It provides two different 
+Vectorizers, one for innermost loops (LoopVectorize) and one for super-word parallelism 
+(SLPVectorize)~\cite{llvmvec}.
+
+This system, however, had quite a few limitations, as the loop vectorizer could only handle
+innermost loops and neither outer loops, complex control flow, nor non-inlined functions. 
+Additionally, while multiple different vectorizations for the same scalar code 
+would be possible, the current vectorizers working directly on the LLVM IR could not model
+and compare the costs of such different vectorization approaches.
+
+With these limitations in mind, Intel started an ongoing refactorization effort to migrate LLVM's
+auto-vectorization pipeline to utilize a more abstract Vectorization Plan 
+(VPlan)~\cite{llvmextloopvec,llvmvplan}. The final goal would be to unite LLVM's auto-vectorization
+in a single flexible system capable of optimizing SLPs, inner, and nested loops with complex 
+control flows. The auto-vectorization pass would create, compare, and transform multiple different 
+VPlans, each modeling a different vectorization approach, and finally materialize the best one into
+the LLVM IR.
+
+LLVM's VPlan is currently being integrated into the existing Loop Vectorizer and is
+already modeling most inner loop vectorizations and transformations. Vectorization for outer 
+loops is also in development and can be enabled by setting the \texttt{-enable-vplan-native-path} 
+flag~\cite{llvmouterloop}. In the future, the plan will be to merge both of these loop vectorization
+paths into one.
 
@@
 \label{sec:background}
-Lorem ipsum
+Modern SIMD ISAs like Intel AVX, ARM NEON or even vector-length agnostic ISAs like ARM SVE provide
+a variety of different instructions to allow vectorization of even complex control flows. To lower
+the complexity for implementing auto-vectorization, the general optimizations are often divided
+into different categories:
+
+\subsection{Inner Loop Vectorization}
+Innermost loops are loops whose body's control flow does not contain any loops, bound or unbound.
+Vectorizing such loops is generall eas
+
+\begin{figure}
+  \centering
+  \includegraphics[width=0.35\textwidth]{images/inner-loop-vec.png}
+  \caption{asdf}
+  \label{fig:inner-loop-vec}
+\end{figure}
+
+\subsection{Super-Word Parallelism (SLP) Vectorization}
+asdf
+
+\subsection{Outer Loop Vectorizaton}
+asdf
+
+\subsection{Function Vectorization}
+asdf
+
+\subsection{Vectorization Constraints}
+asdf
 

### 2024-11-20 paper/paper.tex
paper

@@
 in a single flexible system capable of optimizing SLPs, inner, and nested loops with complex 
-control flows. The auto-vectorization pass would create, compare, and transform multiple different 
-VPlans, each modeling a different vectorization approach, and finally materialize the best one into
-the LLVM IR.
+control flows. The auto-vectorization pass would create and transform multiple different 
+VPlans, each modeling a different vectorization approach, abstracted away from the underlying LLVM IR.
+Those VPlans are compared against each other based on a cost model and finally the best one, which
+might also be not to vectorize at all, is materialized into the LLVM IR.
 
 LLVM's VPlan is currently being integrated into the existing Loop Vectorizer and is
-already modeling most inner loop vectorizations and transformations. Vectorization for outer 
-loops is also in development and can be enabled by setting the \texttt{-enable-vplan-native-path} 
-flag~\cite{llvmouterloop}. In the future, the plan will be to merge both of these loop vectorization
-paths into one.
+already modeling most inner loop vectorizations and transformations~\cite{llvmvplanupdate}. 
+Vectorization for outer loops is also in development and can be enabled by setting 
+the \texttt{-enable-vplan-native-path} flag~\cite{llvmouterloop}. In the future, the plan will 
+be to merge both of these loop vectorization paths into one.
 
@@
 a variety of different instructions to allow vectorization of even complex control flows. To lower
-the complexity for implementing auto-vectorization, the general optimizations are often divided
+the complexity for implementing auto-vectorization, the general optimizations are often seperated
 into different categories:
@@
 \subsection{Inner Loop Vectorization}
-Innermost loops are loops whose body's control flow does not contain any loops, bound or unbound.
-Vectorizing such loops is generall eas
+Innermost loops are loops whose body's control flow does not contain any more loops, bound or unbound.
+Vectorization of these loops are simpler in comparison as the only option to vectorize is over the
+loop's induction variable.
+
+There are various ways to transform the control flow of a loop to accomodate operating on widened
+vector registers. Figure \ref{fig:inner-loop-vec} shows an example of an if-conversion. Some more
+examples can also be found in LLVM's Auto-Vectorization Documentation~\cite{llvmvec}.
 
@@
   \includegraphics[width=0.35\textwidth]{images/inner-loop-vec.png}
-  \caption{asdf}
+  \caption{A code snippet of a simple loop that would be vectorized over the variable \texttt{i}. 
+  The conditional branch would be flattened into a masked-addition. The mask would be generated 
+  based on \texttt{z[i]}.}
   \label{fig:inner-loop-vec}
@@
 \subsection{Super-Word Parallelism (SLP) Vectorization}
-asdf
+SLP Vectorization vectorizes similar independent instructions. In the context of auto-vectorization a 
+compiler might have, for example, fully unrolled a smaller loop in a previous simplifcation pass. 
+The resulting similar scalar instructions might then be vectorized later on by an SLP vectorizer.
 
 \subsection{Outer Loop Vectorizaton}
-asdf
+Outer loop vectorization focuses on control flow with nested loops. Such loops often present
+more challenges for auto-vectorization, as one could possibly vectorize over the outer loop,
+the inner loop or a mix of both, representing a hybrid/multi-dimensional vectorization approach. Carrying out and
+comparing costs of such different vectorization approaches/transformations can be complex and
+require a flexible system like VPlans.
+
+Figure \ref{fig:outer-loop-vec} shows a nested loop code-snippet where vectorization over the
+outer loop would be beneficial.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=0.39\textwidth]{images/outer-loop-vec.png}
+  \caption{A code snippet of a nested loop where it would be beneficial to vectorize over the 
+  outer loop. As the inner loop has a too low trip count (6) to vectorize and additionally
+  the memory access pattern of \texttt{y} would be adjacent compared to scattered when
+  vectorizing over \texttt{i} instead of \texttt{j}.}
+  \label{fig:outer-loop-vec}
+\end{figure}
 
 \subsection{Function Vectorization}
-asdf
+Function vectorization refers to the vectorization of full functions. A loop might contain a
+function call, that can not be inlined, passing data related to the iteration variable. During
+auto-vectorization this call would possibly be replaced with a call to a newly constructed function 
+with the same semantic control flow but operating on multiple values at once through vector
+registers instead.
 
-\subsection{Vectorization Constraints}
-asdf
+\subsection{Vectorization Similarities}
+All of the above mentioned vectorization categories can be brought into relatation with each other. 
+For example function vectorization can be transformed into loop vectorization by creating a loop 
+around the function body and vectorizing it and vice versa. Similar is true for SLP Vectorization 
+and loop vectorization by unrolling a loop.
+
+\subsection{Vectorization Legality}
+Not all loops can be vectorized. There are various factors that can completely hinder any
+vectorization attempts.
+
+% TODO
+
+\subsection{Vectorization Costs}
+While it may just not be legal to vectorize a given control flow, sometimes the vectorized code 
+might also just perform worse than its scalar equivalent.
+
+Some control flow transformations that seem beneficial at compile time might reveal a performance
+bottleneck at runtime. Considering the code snippet from Figure \ref{fig:inner-loop-vec}, it might
+be that at runtime almost all of \texttt{z[i]} evaluate to false. The branch predictor of a CPU
+would then be able to guess the correct branch in the scalar code. The vectorized version would
+always load the \texttt{y[i]} into memory only for the addition to be masked away. There are
+different approaches to avoid such runtime penalties such as branch-on-superword-condition code 
+(BOSCC)~\cite{10.5555/1299042.1299055,llvmboscc} or active-lane 
+consolidation (ALC)~\cite{10.1007/s11227-022-04359-w,10.5555/3615924.3615932}.
+
+Some vector instructions can be quite costly on most CPUs. An example would be gather/scatter
+or shuffeling and permutation instructions. Control flow with a lot of non-adjacent 
+memory load/stores or operations that combine different vector elements such as reductions
+can decrease vectorization performance.
+
+Additionally the vectorized code size will always be larger than the scalar equivalent. This is
+due to many factors: (1) The transformed control flow oftentimes simply includes more instructions.
+(2) As the loop trip count might not be a perfect multiple of the vectorization factor, a scalar
+epilogue/remainder of the loop might be used to handle the tail iterations. (3) Various runtime
+checks might be inserted throughout the vectorized control flow to handle
 

### 2024-11-20 paper/paper.tex
paper

@@
 
-However, as manual code vectorization can quickly become very time-consuming, especially when
-supporting different CPU architectures, modern compilers aim to automatically transform scalar code
+However, manual code vectorization can quickly become time-consuming, especially when
+supporting different CPU architectures. Modern compilers aim to automatically transform scalar code
 to use vectorization when applicable.
@@
 VPlans, each modeling a different vectorization approach, abstracted away from the underlying LLVM IR.
-Those VPlans are compared against each other based on a cost model and finally the best one, which
-might also be not to vectorize at all, is materialized into the LLVM IR.
+Those VPlans are compared against each other based on a cost model, and finally, the best one, which
+might also not vectorize at all, is materialized into the LLVM IR.
 
@@
 \label{sec:background}
-Modern SIMD ISAs like Intel AVX, ARM NEON or even vector-length agnostic ISAs like ARM SVE provide
-a variety of different instructions to allow vectorization of even complex control flows. To lower
-the complexity for implementing auto-vectorization, the general optimizations are often seperated
-into different categories:
+In general, vectorization techniques can be separated into different categories:
 
@@
 Innermost loops are loops whose body's control flow does not contain any more loops, bound or unbound.
-Vectorization of these loops are simpler in comparison as the only option to vectorize is over the
-loop's induction variable.
+Vectorization of these loops is more straightforward as the only option to vectorize is over 
+the loop's induction variable.
 
-There are various ways to transform the control flow of a loop to accomodate operating on widened
+There are various ways to transform the control flow of a loop to accommodate operating on widened
 vector registers. Figure \ref{fig:inner-loop-vec} shows an example of an if-conversion. Some more
@@
   \caption{A code snippet of a simple loop that would be vectorized over the variable \texttt{i}. 
-  The conditional branch would be flattened into a masked-addition. The mask would be generated 
+  The conditional branch would be flattened into a masked addition. The mask would be generated 
   based on \texttt{z[i]}.}
@@
 \subsection{Super-Word Parallelism (SLP) Vectorization}
-SLP Vectorization vectorizes similar independent instructions. In the context of auto-vectorization a 
-compiler might have, for example, fully unrolled a smaller loop in a previous simplifcation pass. 
-The resulting similar scalar instructions might then be vectorized later on by an SLP vectorizer.
+SLP Vectorization vectorizes similar independent instructions. A compiler might have, for example, 
+fully unrolled a smaller loop in a previous code simplification pass. The resulting similar scalar 
+instructions might then be vectorized later by an SLP vectorizer.
 
@@
 Outer loop vectorization focuses on control flow with nested loops. Such loops often present
-more challenges for auto-vectorization, as one could possibly vectorize over the outer loop,
-the inner loop or a mix of both, representing a hybrid/multi-dimensional vectorization approach. Carrying out and
-comparing costs of such different vectorization approaches/transformations can be complex and
-require a flexible system like VPlans.
+more challenges for auto-vectorization, as one could vectorize over the outer loop,
+inner loop, or a mix of both, representing a hybrid/multi-dimensional vectorization approach. 
+Carrying out and comparing costs of such different vectorization approaches can be complex 
+and require a flexible system like, for example, VPlans.
 
-Figure \ref{fig:outer-loop-vec} shows a nested loop code-snippet where vectorization over the
+Figure \ref{fig:outer-loop-vec} shows a nested loop code snippet where vectorization over the
 outer loop would be beneficial.
@@
   \caption{A code snippet of a nested loop where it would be beneficial to vectorize over the 
-  outer loop. As the inner loop has a too low trip count (6) to vectorize and additionally
-  the memory access pattern of \texttt{y} would be adjacent compared to scattered when
+  outer loop. The inner loop has a trip count that is too low (6) to vectorize. Additionally,
+  the memory access pattern of \texttt{y} would be consecutive compared to scattered when
   vectorizing over \texttt{i} instead of \texttt{j}.}
@@
 \subsection{Function Vectorization}
-Function vectorization refers to the vectorization of full functions. A loop might contain a
-function call, that can not be inlined, passing data related to the iteration variable. During
-auto-vectorization this call would possibly be replaced with a call to a newly constructed function 
-with the same semantic control flow but operating on multiple values at once through vector
-registers instead.
+Function vectorization refers to the vectorization of entire functions. A loop might contain a
+non-inlined function call, passing data related to the iteration variable. During auto-vectorization, 
+this call could be replaced with a call to a newly constructed function with the same semantic 
+control flow but operating on multiple values at once through vector registers instead.
 
 \subsection{Vectorization Similarities}
-All of the above mentioned vectorization categories can be brought into relatation with each other. 
-For example function vectorization can be transformed into loop vectorization by creating a loop 
-around the function body and vectorizing it and vice versa. Similar is true for SLP Vectorization 
-and loop vectorization by unrolling a loop.
+All of the above-mentioned vectorization techniques can be brought into relation with each other. 
+For example, function vectorization can be transformed into loop vectorization by creating a loop 
+around the function body and vectorizing it and vice versa. The same applies for SLP Vectorization 
+and loop vectorization when unrolling a loop.
 
@@
 epilogue/remainder of the loop might be used to handle the tail iterations. (3) Various runtime
-checks might be inserted throughout the vectorized control flow to handle
+checks might be inserted throughout the vectorized control flow to handle % TODO
 

### 2024-11-20 paper/paper.tex
paper

@@
 There are various ways to transform the control flow of a loop to accommodate operating on widened
-vector registers. Figure \ref{fig:inner-loop-vec} shows an example of an if-conversion. Some more
+vector registers. Figure~\ref{fig:inner-loop-vec} shows an example of an if-conversion. Some more
 examples can also be found in LLVM's Auto-Vectorization Documentation~\cite{llvmvec}.
@@
 
-Figure \ref{fig:outer-loop-vec} shows a nested loop code snippet where vectorization over the
+Figure~\ref{fig:outer-loop-vec} shows a nested loop code snippet where vectorization over the
 outer loop would be beneficial.
@@
 \subsection{Vectorization Legality}
-Not all loops can be vectorized. There are various factors that can completely hinder any
+Not all loops can be vectorized. There are various factors that can ultimately hinder any
 vectorization attempts.
 
-% TODO
+Some of the most common vectorization barriers are data dependencies. As the vectorized version
+of a code executes for $N$ (Vectorization Factor) data streams simultaneously, it inherently changes
+the order of operations. To guarantee that the semantics of the code do not change, all
+operations on these $N$ data streams must, generally, be independent. The simplest case
+where this is not provided is when the $i$-th loop iteration depends on any of the $(i-N)$-th 
+previous iterations.
+
+While explicit data dependencies are easier to see, even a simple loop, as in 
+Figure~\ref{fig:inner-loop-vec}, could carry an implicit dependency in line 3, if
+the \texttt{x} and \texttt{y} point to overlapping memory regions (pointer aliasing).
+
+Not all data dependencies directly hinder any vectorization. One exception can 
+sometimes be made for reduction idioms, as seen in the inner loop in 
+Figure~\ref{fig:outer-loop-vec}, where \texttt{sum} would be a reduction variable.
+
+Additionally, the control flow should be primarily straight-line code. While most if-else 
+constructs can be flattened to always execute both branches along with masking, the same is not 
+valid for switch-case statements or too complex control flows.
 
 \subsection{Vectorization Costs}
-While it may just not be legal to vectorize a given control flow, sometimes the vectorized code 
-might also just perform worse than its scalar equivalent.
+While it may not be legal to vectorize a given control flow, the vectorized code sometimes 
+performs worse than its scalar equivalent.
 
 Some control flow transformations that seem beneficial at compile time might reveal a performance
-bottleneck at runtime. Considering the code snippet from Figure \ref{fig:inner-loop-vec}, it might
-be that at runtime almost all of \texttt{z[i]} evaluate to false. The branch predictor of a CPU
+bottleneck at runtime. Considering the code snippet from Figure~\ref{fig:inner-loop-vec}, it might
+be that almost all \texttt{z[i]} evaluate to false at runtime. The branch predictor of a CPU
 would then be able to guess the correct branch in the scalar code. The vectorized version would
-always load the \texttt{y[i]} into memory only for the addition to be masked away. There are
-different approaches to avoid such runtime penalties such as branch-on-superword-condition code 
+always load the \texttt{y[i]} into memory only for the addition to be masked away. Different 
+approaches to avoid such runtime penalties are branch-on-superword-condition code 
 (BOSCC)~\cite{10.5555/1299042.1299055,llvmboscc} or active-lane 
@@
 
-Some vector instructions can be quite costly on most CPUs. An example would be gather/scatter
-or shuffeling and permutation instructions. Control flow with a lot of non-adjacent 
-memory load/stores or operations that combine different vector elements such as reductions
-can decrease vectorization performance.
+Some vector instructions can be costly on most CPUs. Examples are gather, scatter, shuffle,
+and permutation instructions. Control flow with many non-adjacent memory loads/stores or 
+operations that combine different vector elements, such as reductions, can decrease vectorization 
+performance.
 
-Additionally the vectorized code size will always be larger than the scalar equivalent. This is
-due to many factors: (1) The transformed control flow oftentimes simply includes more instructions.
+Additionally, the vectorized code size will always be larger than the scalar equivalent. This is
+due to many factors: (1) The transformed control flow often includes more instructions.
 (2) As the loop trip count might not be a perfect multiple of the vectorization factor, a scalar
 epilogue/remainder of the loop might be used to handle the tail iterations. (3) Various runtime
-checks might be inserted throughout the vectorized control flow to handle % TODO
+checks might be inserted throughout the vectorized control flow to handle, for example, pointers 
+to overlapping memory regions, in which case the scalar version of the loop is executed.
 

### 2024-11-20 paper/paper.tex
paper

@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 the \texttt{-enable-vplan-native-path} flag~\cite{llvmouterloop}. In the future, the plan will 
-be to merge both of these loop vectorization paths into one.
+be to merge both of these loop vectorization paths into one. Merging the existing SLP Vectorizer
+with VPlan is also planned.
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 \label{sec:vplan}
-Lorem ipsum
+Originally LLVM implements two different vectorizers for auto-vectorization, LoopVectorize (LV)
+and SLPVectorize. These vectorizers are a part of LLVM's middle-end and run after module 
+simplification. As VPlan has not yet been integrated into SLP-Vectorization we will only focus on
+explaining the original LoopVectorize structure.
 
+LoopVectorize was designed to focus only on vectorization of innermost loops~\cite{llvmintrvplan}. 
+When starting with a initial scalable loop in IR it will first run a legality phase to check 
+whetever and how it is legal to vectorize the loop. This phase would produce artifacts like if 
+and what runtime aliasing checks would be required. If some vectorization would be legal the 
+cost model would run next, creating more artifacts based on the scalar IR and legality artifacts. 
+These cost model artifacts would associate different vectorization possibilities with a fixed cost.
+Lastly the transformation phase would vectorize the underlying IR based on all the produced 
+artifacts.
+
+However, this LoopVectorize architecture had several problems. For once extending the rigid system
+to include outer loop and nested loop vectorizations would have been very challenging. Producing
+and considering more and more artifacts when trying to transform the scalar into a vectorized
+loop all at once does not scale well.
+
+\subsection{State of Vectorization Plan (VPlan)}
+
+\subsection{Future of VPlan}
+%\subsection{Integration with LV}
+%\subsection{Outer Loop Vectorization}
+%\subsection{Merging with SLP Vectorize}
+%\subsection{Function Vectorization}
+
+\begin{figure}
+  \centering
+  \includegraphics[width=0.475\textwidth]{images/vplan-future.png}
+  \caption{The planned future architecture for VPlan. The VPlan optimization would
+  work as iteratievely applying VPlan-to-VPlan transformations.}
+  \label{fig:vplan-future}
+\end{figure}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2024-11-21 paper/paper.tex
paper

@@
 Some of the most common vectorization barriers are data dependencies. As the vectorized version
-of a code executes for $N$ (Vectorization Factor) data streams simultaneously, it inherently changes
+of a code executes for Vectorization Factor (VF) $N$ data streams simultaneously, it inherently changes
 the order of operations. To guarantee that the semantics of the code do not change, all
@@
 When starting with a initial scalable loop in IR it will first run a legality phase to check 
-whetever and how it is legal to vectorize the loop. This phase would produce artifacts like if 
+whetever and how it is legal to vectorize the loop. This phase would produce artifacts/maps like if 
 and what runtime aliasing checks would be required. If some vectorization would be legal the 
 cost model would run next, creating more artifacts based on the scalar IR and legality artifacts. 
-These cost model artifacts would associate different vectorization possibilities with a fixed cost.
+These cost model artifacts would associate different vectorization factors with a cost.
 Lastly the transformation phase would vectorize the underlying IR based on all the produced 
@@
 
-However, this LoopVectorize architecture had several problems. For once extending the rigid system
-to include outer loop and nested loop vectorizations would have been very challenging. Producing
-and considering more and more artifacts when trying to transform the scalar into a vectorized
-loop all at once does not scale well.
+However, this rigid system of LoopVectorize was a problem. Extending LV to include outer loop 
+and nested loop vectorizations would have been very challenging. Producing and considering more and 
+more artifacts when trying to transform the scalar loop into a vectorized one all at once did not 
+scale well.
 
-\subsection{State of Vectorization Plan (VPlan)}
+As a potential solution Intel proposed a long-term effort to refactor LLVM's whole auto-vectorization
+using a system called Vectorization Plan (VPlan)~\cite{llvmextloopvec}.
 
-\subsection{Future of VPlan}
-%\subsection{Integration with LV}
-%\subsection{Outer Loop Vectorization}
-%\subsection{Merging with SLP Vectorize}
-%\subsection{Function Vectorization}
+The goal was to unite any auto-vectorization systems (SLP, Inner-/Outer-Loop) into a single flexible
+system~\cite{llvmintrvplan,llvmvplanstate}, as seen in Figure~\ref{fig:vplan-future}. This system would use VPlans to represent different 
+vectorization approaches, abstracted away from the underlying IR. Each VPlan would be able to 
+calculate its cost and transform itself to modify the final IR. At first the Legality 
+phase (yellow) would check if the scalar code would be legal to vectorize at all. If so it would 
+create different initial VPlans, representing the different initial ways to vectorize. In the 
+following Planning phase (magenta) these initial VPlans would be optimized using various 
+VPlan-to-VPlan transformations. Those transformations would be supervised by the VPlan's cost model.
+Finally, the best VPlan would be chosen and materialized/executed to modify the underlying IR.
+The best VPlan might also be to not vectorize at all (VF $=1$).
 
@@
   \caption{The planned future architecture for VPlan. The VPlan optimization would
-  work as iteratievely applying VPlan-to-VPlan transformations.}
+  work as iteratively applying VPlan-to-VPlan transformations.}
   \label{fig:vplan-future}
 \end{figure}
+
+\subsection{Current State of VPlan}
+
+\subsection{Future of VPlan}
+Many of the VPlan internals and systems surrounding it are still in active development. To stay
+up-to-date with the latest developments regarding auto-vectorization in LLVM one can follow the 
+Community Forum\footnote{\url{https://discourse.llvm.org/c/ir-optimizations/loop-optimizations/62}} or
+attend the monthly online 
+meetings\footnote{\url{https://discourse.llvm.org/t/monthly-vectorizer-online-sync-up/78978/3}}.
+
+\subsubsection{VPlan and Integration with LV}
+The main effort of setting up the VPlan structure and integrating it first into LV has seen
+tremendous effort over the last years.
+
+\subsubsection{Outer Loop Vectorization}
+Development around outer loop vectorization has slowed down a bit as focus shifted more on VPlan
+internals and integration into LV.
+
+\subsubsection{Merging with SLP Vectorize}
+Merging the existing SLP Vectorizer with the current VPlan infrastructure is in the early phase
+of planning.
+
+\subsubsection{Function Vectorization}
+As far as we now there is currently no active development for function vectorization in the
+llvm-project's upstream yet.
 

### 2024-11-22 code/simple_loop.c
add inner loop conditional ir code

@@ -15,7 +15,7 @@
 //     }
 // }
 
-void inner_loop_conditional(const int N, float* x, const float* y, const bool* z) {
+void inner_loop_conditional(const int N, float* x, const float* y, const int* z) {
     for (int i = 0; i < N; i++) {
         if (z[i]) {
             x[i] += y[i];

### 2024-11-22 paper/paper.tex
paper

@@
 calculate its cost and transform itself to modify the final IR. At first the Legality 
-phase (yellow) would check if the scalar code would be legal to vectorize at all. If so it would 
-create different initial VPlans, representing the different initial ways to vectorize. In the 
-following Planning phase (magenta) these initial VPlans would be optimized using various 
-VPlan-to-VPlan transformations. Those transformations would be supervised by the VPlan's cost model.
-Finally, the best VPlan would be chosen and materialized/executed to modify the underlying IR.
-The best VPlan might also be to not vectorize at all (VF $=1$).
+phase (yellow) would check if the scalar code would be legal to vectorize at all and if so initial 
+VPlans would be constructed. In the following Planning phase (magenta) these initial VPlans would 
+be optimized using various VPlan-to-VPlan transformations. Those transformations would be supervised 
+by the VPlan's cost model. Finally, the best VPlan would be chosen and materialized/executed, 
+modifying the underlying IR. The best VPlan might also be to not vectorize at all (VF $=1$).
 
@@
 
+In its current state (LLVM 19.1.4) VPlan is deeply integrated into LV.
+
+\begin{figure*}[ht!]
+  \centering
+  \includegraphics[width=\textwidth]{images/inner-loop-vplan-hcfg.png}
+  \caption{Hierarchical CFG of a VPlan modelling the vectorized version of the loop in 
+  Figure~\ref{fig:inner-loop-vec}.}
+  \label{fig:inner-loop-vplan-hcfg}
+\end{figure*}
+
 \subsection{Future of VPlan}
@@
 
-\subsubsection{VPlan and Integration with LV}
+\subsubsection{General VPlan and Integration with LV}
 The main effort of setting up the VPlan structure and integrating it first into LV has seen
-tremendous effort over the last years.
+tremendous effort over the last years. Florian Hahn presented future directions for 
+VPlan~\cite{llvmvplanupdate}, which mostly evolve around transforming more LV-based decisions into
+explicit VPlan-to-VPlan transformations. Additionally, a lot of cost model related decisions 
+still end up being influenced by the old legacy LV cost model, however, transition to a VPlan-based 
+cost model has already started.
 
 \subsubsection{Outer Loop Vectorization}
-Development around outer loop vectorization has slowed down a bit as focus shifted more on VPlan
-internals and integration into LV.
+Development around outer loop vectorization has slowed down a bit as focus shifted more onto VPlan
+internals and integration into LV. The inner loop path and outer loop path are currently still
+seperated by the experimental \texttt{-enable-vplan-native-path} flag. Further developing the 
+outer loop path and eventually merging it with the innermost loop path is planned.
 

### 2024-11-22 paper/paper.tex
paper

@@
 \label{sec:vplan}
-Originally LLVM implements two different vectorizers for auto-vectorization, LoopVectorize (LV)
+
+\subsection{History}
+LLVM implements two different vectorizers for auto-vectorization, LoopVectorize (LV)
 and SLPVectorize. These vectorizers are a part of LLVM's middle-end and run after module 
-simplification. As VPlan has not yet been integrated into SLP-Vectorization we will only focus on
-explaining the original LoopVectorize structure.
+simplification has taken place. As VPlan has not yet been integrated into SLP-Vectorization 
+we will only focus on explaining the original LoopVectorize structure.
 
 LoopVectorize was designed to focus only on vectorization of innermost loops~\cite{llvmintrvplan}. 
-When starting with a initial scalable loop in IR it will first run a legality phase to check 
+When starting with a initial scalable loop in IR it would first run a legality phase to check 
 whetever and how it is legal to vectorize the loop. This phase would produce artifacts/maps like if 
-and what runtime aliasing checks would be required. If some vectorization would be legal the 
-cost model would run next, creating more artifacts based on the scalar IR and legality artifacts. 
-These cost model artifacts would associate different vectorization factors with a cost.
-Lastly the transformation phase would vectorize the underlying IR based on all the produced 
-artifacts.
+and what runtime aliasing checks would be required. Then the cost model ran next, creating more 
+artifacts based on the scalar IR and legality artifacts. These cost model artifacts would 
+associate different vectorization factors with a cost. Lastly the transformation phase would 
+vectorize the underlying IR in one go based on all the produced artifacts.
 
-However, this rigid system of LoopVectorize was a problem. Extending LV to include outer loop 
-and nested loop vectorizations would have been very challenging. Producing and considering more and 
-more artifacts when trying to transform the scalar loop into a vectorized one all at once did not 
-scale well.
+However, this rigid flow of LoopVectorize was a problem. Extending LV to include outer loop 
+and nested loop vectorizations would have been very challenging, as LV was designed to flatten a
+loops control flow into a single basic block. Producing and considering more and 
+more artifacts when trying to transform the scalar loop into a vectorized one all at once did also 
+not scale well.
 
@@
 \subsubsection{Function Vectorization}
-As far as we now there is currently no active development for function vectorization in the
+As far as we now there is currently no active development for whole-function vectorization in the
 llvm-project's upstream yet.

### 2024-11-22 paper/paper.tex
paper

@@
 
-In its current state (LLVM 19.1.4) VPlan is deeply integrated into LV.
+In its current state (LLVM 19.1.4) VPlan is deeply integrated into LV. To better understand the
+current structure of a VPlan, modelling a certain vectorization approach, we will use the
+simple conditional loop from Figure~\ref{fig:inner-loop-vec} as an example.
 
-\begin{figure*}[ht!]
+%TODO: Combine inner-loop-scalar-ir figure and inner-loop-vplan-hcfg into a single
+%       double column figure.
+\begin{figure}
   \centering
-  \includegraphics[width=\textwidth]{images/inner-loop-vplan-hcfg.png}
-  \caption{Hierarchical CFG of a VPlan modelling the vectorized version of the loop in 
-  Figure~\ref{fig:inner-loop-vec}.}
+  \includegraphics[width=0.475\textwidth]{images/inner-loop-scalar-loop-ir.png}
+  \caption{The scalar IR for the main loop part of the conditional loop depicted in
+  Figure~\ref{fig:inner-loop-vec}. Note that \texttt{getelementptr inbounds}
+  was abbreviated to \texttt{getelmptr inbs}.}
+  \label{fig:inner-loop-scalar-ir}
+\end{figure}
+
+\begin{figure}
+  \centering
+  \includegraphics[width=0.475\textwidth]{images/inner-loop-vplan-hcfg-loop-body-2.png}
+  \caption{The vectorized loop part of a hierarchical CFG of a VPlan modelling the conditional loop
+  depicted in Figure~\ref{fig:inner-loop-vec}. This VPlan has a VF of 4. Note that some keywords
+  are abbreviated: \texttt{arrayidx} $\equiv$ \texttt{arridx}; 
+  \texttt{getelementptr inbounds} $\equiv$ \texttt{getelmptr inbs}}
   \label{fig:inner-loop-vplan-hcfg}
-\end{figure*}
+\end{figure}
 
@@
 \label{sec:summary}
-Lorem ipsum
+In conclusion, VPlan has been in active development for several years now. When compiling with
+LLVM VPlan is used to model initial vectorization approaches, keep track of different
+optimization decisions and finally execute the best VPlan to transform IR. 
+However, there are still some VPlan-to-VPlan transformations and a VPlan-based cost model missing, 
+before the legacy LV code can be completely discarded.
+
+Outer loop vectorization support for VPlan is still experimental, seperated from the inner loop
+LV integration, and runs on a divergent experimental path (\texttt{-enable-vplan-native-path}).
+In the near future development on outer loop vectorization will continue and merging it with LV's
+VPlan path is planned.
+
+Active development of other approaches such as SLP and whole-function vectorization based on the 
+VPlan architecture has not been started yet.
 
@@
 
+\appendix
+
+\section{Appendix: Additional Figures}
+
+\begin{figure*}
+  \centering
+  \includegraphics[width=\textwidth]{images/inner-loop-vplan-hcfg.png}
+  \caption{Full hierarchical CFG of a VPlan modelling the vectorized version of the loop 
+  depicted in Figure~\ref{fig:inner-loop-vec} as of LLVM 19.1.4.}
+  \label{fig:inner-loop-vplan-hcfg}
+\end{figure*}
+
 \end{document}

### 2024-11-22 paper/paper.tex
paper

@@
 Vectorization for outer loops is also in development and can be enabled by setting 
-the \texttt{-enable-vplan-native-path} flag~\cite{llvmouterloop}. In the future, the plan will 
-be to merge both of these loop vectorization paths into one. Merging the existing SLP Vectorizer
-with VPlan is also planned.
+the \texttt{-enable-vplan-native-path} feature flag~\cite{llvmouterloop,llvmouterloopstatus}. 
+In the future, the plan will be to merge both of these loop vectorization paths into one. 
+Merging the existing SLP Vectorizer with VPlan is also planned.
 
@@
 
-\subsection{Current State of VPlan}
+\subsection{State of VPlan (LLVM 19.1.4)}
 
-In its current state (LLVM 19.1.4) VPlan is deeply integrated into LV. To better understand the
-current structure of a VPlan, modelling a certain vectorization approach, we will use the
+In its current state VPlan is deeply integrated into LV.
+When auto-vectorizing, LV starts by checking all the different loops in the IR.
+If it finds one, it will try to vectorize it\\(\texttt{LoopVectorizePass::processLoop}).
+Several factors might hinder vectorization for a given loop: 
+(1) If the loop was already vectorized before, it won't be vectorized again. 
+(2) The Legality phase checks if an how it is legal to vectorize the scalar
+loop. (3) The loop's trip count might be too small to make vectorization profitable. (4) The cost
+of runtime checks for e.g. pointer aliasing can be too high.
+
+The initial VPlans are constructed based on the legality checks and the cost decisions the legacy
+LV cost model made referencing the scalar IR\\(\texttt{LoopVectorizationPlanner::plan}). Then the
+optimal Vectorization Factor (VF) and Interleave Count (IC) are picked. 
+The best VPlan is picked based on the optimal VF. A VF of 1 means it is not
+profitable to vectorize the loop, however, interleaving/unrolling can still end up modifying the
+underlying IR.
+
+The best VPlan is then executed\\(\texttt{LoopVectorizationPlanner::executePlan}). In case of
+vectorization (VF $>1$), LV will also directly try to vectorize the epilogue loop.
+
+\subsubsection{VPlan Structure}
+To better understand the
+underlying structure of a VPlan, modelling a certain vectorization approach, we will use the
 simple conditional loop from Figure~\ref{fig:inner-loop-vec} as an example.
+
+Figure~\ref{fig:inner-loop-scalar-ir} shows the IR of the main loop basic blocks \texttt{for.body},
+\texttt{if.then}, and \texttt{for.inc}. Figure~\ref{fig:inner-loop-vplan-hcfg} shows the equivalent
+hierarchical control-flow-graph (CFG) of a VPlan modelling a vectorization approach for this 
+scalar loop.
 
@@
 
-\subsubsection{General VPlan and Integration with LV}
+\subsubsection{VPlan and Integration with LV}
 The main effort of setting up the VPlan structure and integrating it first into LV has seen
@@
 
-Outer loop vectorization support for VPlan is still experimental, seperated from the inner loop
+Outer loop vectorization support via VPlan is still experimental, seperated from the inner loop
 LV integration, and runs on a divergent experimental path (\texttt{-enable-vplan-native-path}).

### 2024-11-23 paper/paper.tex
paper

@@
 Vectorization for outer loops is also in development and can be enabled by setting 
-the \texttt{-enable-vplan-native-path} feature flag~\cite{llvmouterloop,llvmouterloopstatus}. 
+the \texttt{-enable-\allowbreak vplan-\allowbreak native-\allowbreak path} feature 
+flag~\cite{llvmouterloop,llvmouterloopstatus}. 
 In the future, the plan will be to merge both of these loop vectorization paths into one. 
@@
 VPlans would be constructed. In the following Planning phase (magenta) these initial VPlans would 
-be optimized using various VPlan-to-VPlan transformations. Those transformations would be supervised 
+be optimized using various VPlan-to-VPlan transformations~\cite{llvmouterloopstatus}. 
+Those transformations would be supervised 
 by the VPlan's cost model. Finally, the best VPlan would be chosen and materialized/executed, 
@@
 \subsection{State of VPlan (LLVM 19.1.4)}
+\emergencystretch 6em
 
@@
 When auto-vectorizing, LV starts by checking all the different loops in the IR.
-If it finds one, it will try to vectorize it\\(\texttt{LoopVectorizePass::processLoop}).
+If it finds one, it will try to vectorize it (\texttt{LoopVectorizePass::\allowbreak processLoop}).
 Several factors might hinder vectorization for a given loop: 
@@
 The initial VPlans are constructed based on the legality checks and the cost decisions the legacy
-LV cost model made referencing the scalar IR\\(\texttt{LoopVectorizationPlanner::plan}). Then the
-optimal Vectorization Factor (VF) and Interleave Count (IC) are picked. 
+LV cost model made referencing the scalar IR (\texttt{LoopVectorizationPlanner::\allowbreak plan}).
+After initial VPlan construction various VPlan-to-VPlan transformations are applied to optimize
+it (\texttt{LoopVectorizationPlanner::\allowbreak buildVPlansWithVPRecipes}).
+
+Then the optimal Vectorization Factor (VF) is calculated based on the set of VPlans and their
+representive cost (\texttt{LoopVectorizationPlanner::\allowbreak cost}), 
+which calculation still involves the legacy LV cost model. 
 The best VPlan is picked based on the optimal VF. A VF of 1 means it is not
@@
 underlying IR.
+Afterwards, the Interleave Count (IC) is determined 
+(\texttt{LoopVectorizationCostModel::\allowbreak selectInterleaveCount}).
 
-The best VPlan is then executed\\(\texttt{LoopVectorizationPlanner::executePlan}). In case of
+The best VPlan is then executed (\texttt{LoopVectorizationPlanner::\allowbreak executePlan}). In case of
 vectorization (VF $>1$), LV will also directly try to vectorize the epilogue loop.
@@
 \subsubsection{VPlan Structure}
-To better understand the
-underlying structure of a VPlan, modelling a certain vectorization approach, we will use the
+To better understand the structure of a VPlan, we will use the
 simple conditional loop from Figure~\ref{fig:inner-loop-vec} as an example.
@@
 Figure~\ref{fig:inner-loop-scalar-ir} shows the IR of the main loop basic blocks \texttt{for.body},
-\texttt{if.then}, and \texttt{for.inc}. Figure~\ref{fig:inner-loop-vplan-hcfg} shows the equivalent
+\texttt{if.then}, and \texttt{for.inc}. Figure~\ref{fig:inner-loop-vplan-hcfg-body} shows the equivalent
 hierarchical control-flow-graph (CFG) of a VPlan modelling a vectorization approach for this 
-scalar loop.
+scalar loop. Other basic blocks like the scalar epilogue loop and it's predication check (is the final
+vector trip count equal to the original trip count) are omitted. See Figure~\ref{fig:inner-loop-vplan-hcfg}
+in Appendix~\ref{appendix-figures} for the fully modeled HCFG.
+
+The hierarchical CFG of a VPlan helps abstract it away from the underlying IR. The nodes of this CFG
+are either \texttt{VPBasicBlocks}, containing a sequence of one or more \texttt{VPRecipes}, or
+\texttt{VPRegionBlocks}, modelling a single-entry-single-exit subgraph.
+In our example the \texttt{vector.body} is a basic block and \texttt{vector loop} is a region block.
+
+\texttt{VPRegionBlocks} can indicate that their contents are to be replicated a certain number of times,
+to model more complex control flow for different VF, UF combinations more efficiently in a single VPlan. 
+In our example \texttt{vector loop} is not to be replicated (\texttt{<x1>}).
+
+Inside our two \texttt{VPBasicBlocks} we have a list of recipes (lines of instructions). A recipe
+models one or more instructions in the executed IR and can be based on one or more instructions.
+from the input IR (ingredients). Note that a recipe might also model a instruction with no reference to any input
+IR (in our example this would be the inversion of our vector mask). There are different types of 
+recipes. \texttt{VPReplicateRecipe} will either \texttt{CLONE} exactly one instruction from the input IR
+into exactly one instruction in the output IR, or \texttt{REPLICATE} an instruction a certain number
+of times into output IR (typically found in replicator \texttt{VPRegionBlocks}). Different 
+\texttt{VPWidenRecipes} exist to transform/\texttt{WIDEN} any input IR instruction into a vectorized 
+version of it. The last remaining recipe type in our example are \texttt{VPInstructions}. They
+\texttt{EMIT} new instructions into the output IR that have no representative in the input IR (no
+ingredients).
+
+Furthermore, VPlan recipes can also model definitions and usages of different values. In our example,
+the comparison mask \texttt{vp<\%6>} is a \texttt{VPValue} that is used by three different 
+\texttt{VPUsers}.
 
@@
   \centering
-  \includegraphics[width=0.475\textwidth]{images/inner-loop-scalar-loop-ir.png}
+  \includegraphics[width=0.475\textwidth]{images/inner-loop-scalar-loop-ir-color.png}
   \caption{The scalar IR for the main loop part of the conditional loop depicted in
@@
   \centering
-  \includegraphics[width=0.475\textwidth]{images/inner-loop-vplan-hcfg-loop-body-2.png}
+  \includegraphics[width=0.475\textwidth]{images/inner-loop-vplan-hcfg-loop-body-color.png}
   \caption{The vectorized loop part of a hierarchical CFG of a VPlan modelling the conditional loop
-  depicted in Figure~\ref{fig:inner-loop-vec}. This VPlan has a VF of 4. Note that some keywords
-  are abbreviated: \texttt{arrayidx} $\equiv$ \texttt{arridx}; 
-  \texttt{getelementptr inbounds} $\equiv$ \texttt{getelmptr inbs}}
-  \label{fig:inner-loop-vplan-hcfg}
+  depicted in Figure~\ref{fig:inner-loop-vec}. Some predefined values are
+  \texttt{vp<\%0>} = VF * UF; \texttt{vp<\%1>} = vector-trip-count; \texttt{vp<\%2>} = original trip-count.}
+  \label{fig:inner-loop-vplan-hcfg-body}
 \end{figure}
@@
 internals and integration into LV. The inner loop path and outer loop path are currently still
-seperated by the experimental \texttt{-enable-vplan-native-path} flag. Further developing the 
+seperated by the experimental \texttt{-enable-\allowbreak vplan-\allowbreak native-\allowbreak path} 
+flag. Further developing the 
 outer loop path and eventually merging it with the innermost loop path is planned.
@@
 Outer loop vectorization support via VPlan is still experimental, seperated from the inner loop
-LV integration, and runs on a divergent experimental path (\texttt{-enable-vplan-native-path}).
+LV integration, and runs on a divergent experimental path (\texttt{-enable-\allowbreak vplan-\allowbreak native-\allowbreak path}).
 In the near future development on outer loop vectorization will continue and merging it with LV's
@@
 
-\section{Appendix: Additional Figures}
+\section{Appendix: Additional Figures}\label{appendix-figures}
 

### 2024-11-23 code/simple_loop.cpp
code

@@ -1,8 +1,8 @@
-#define ENABLE_VECTORIZATION 0
+#define ENABLE_VECTORIZATION 1
 
 #if ENABLE_VECTORIZATION
 #define VECTORIZABLE_LOOP \
-    _Pragma("clang loop vectorize(enable) interleave(enable) vectorize_predicate(disable)")
+    _Pragma("clang loop vectorize_width(4) interleave_count(1)")
 #else
 #define VECTORIZABLE_LOOP \
     _Pragma("clang loop vectorize(disable) interleave(disable)")
@@ -14,11 +14,19 @@
 //         x[i] = y[i] + z[i];
 //     }
 // }
+//
+// void inner_loop_conditional(const int N, float* x, const float* y, const bool* z) {
+//     for (int i = 0; i < N; i++) {
+//         if (z[i]) {
+//             x[i] += y[i];
+//         }
+//     }
+// }
 
-void inner_loop_conditional(const int N, float* x, const float* y, const bool* z) {
+void inner_loop_conditional_intdiv(const int N, int *a, int b, int *c) {
     for (int i = 0; i < N; i++) {
-        if (z[i]) {
-            x[i] += y[i];
+        if (a[i] > 1337) {
+            a[i] = c[i] / b;
         }
     }
 }

### 2024-11-23 paper/paper.tex
paper

@@
 
-Figure~\ref{fig:inner-loop-scalar-ir} shows the IR of the main loop basic blocks \texttt{for.body},
+Figure~\ref{fig:inner-loop-scalar-ir} shows the scalar IR of the main loop basic blocks \texttt{for.body},
 \texttt{if.then}, and \texttt{for.inc}. Figure~\ref{fig:inner-loop-vplan-hcfg-body} shows the equivalent
@@
 in Appendix~\ref{appendix-figures} for the fully modeled HCFG.
+
+%TODO: Combine inner-loop-scalar-ir figure and inner-loop-vplan-hcfg into a single
+%       double column figure.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=0.475\textwidth]{images/inner-loop-scalar-loop-ir-color.png}
+  \caption{The scalar IR for the main loop part of the conditional loop depicted in
+  Figure~\ref{fig:inner-loop-vec}. Note that \texttt{getelementptr inbounds}
+  was abbreviated to \texttt{getelmptr inbs}.}
+  \label{fig:inner-loop-scalar-ir}
+\end{figure}
+
+\begin{figure}
+  \centering
+  \includegraphics[width=0.475\textwidth]{images/inner-loop-vplan-hcfg-loop-body-color.png}
+  \caption{The vectorized loop part of a hierarchical CFG of a VPlan modelling the conditional loop
+  depicted in Figure~\ref{fig:inner-loop-vec}. Some predefined values are
+  \texttt{vp<\%0>} = VF * UF; \texttt{vp<\%1>} = vector-trip-count; \texttt{vp<\%2>} = original trip-count.}
+  \label{fig:inner-loop-vplan-hcfg-body}
+\end{figure}
+
+The different colored parts in the scalar IR can be found with the same color in the modeled VPlan.
+The only completely new part that VPlan models in this example is the handling of the scalar epilogue
+loop (purple).
 
@@
 \texttt{VPUsers}.
-
-%TODO: Combine inner-loop-scalar-ir figure and inner-loop-vplan-hcfg into a single
-%       double column figure.
-\begin{figure}
-  \centering
-  \includegraphics[width=0.475\textwidth]{images/inner-loop-scalar-loop-ir-color.png}
-  \caption{The scalar IR for the main loop part of the conditional loop depicted in
-  Figure~\ref{fig:inner-loop-vec}. Note that \texttt{getelementptr inbounds}
-  was abbreviated to \texttt{getelmptr inbs}.}
-  \label{fig:inner-loop-scalar-ir}
-\end{figure}
-
-\begin{figure}
-  \centering
-  \includegraphics[width=0.475\textwidth]{images/inner-loop-vplan-hcfg-loop-body-color.png}
-  \caption{The vectorized loop part of a hierarchical CFG of a VPlan modelling the conditional loop
-  depicted in Figure~\ref{fig:inner-loop-vec}. Some predefined values are
-  \texttt{vp<\%0>} = VF * UF; \texttt{vp<\%1>} = vector-trip-count; \texttt{vp<\%2>} = original trip-count.}
-  \label{fig:inner-loop-vplan-hcfg-body}
-\end{figure}
 

### 2024-11-23 paper/outline.tex
paper

@@
 \begin{abstract}
-Lorem ipsum.
+abstract
 \end{abstract}
@@
   Vectorization Plan $\rightarrow$ ongoing 8 year long effort started by Intel
-  \item rough paper outline
 \end{enumerate}
@@
 \begin{enumerate}
-  \item Phase 1: Legality
-  \item Phase 2: Planning (Vectorization Plans)
-  \begin{enumerate}
-    \item Building initial VPlan based on Phase 1 Constraints (Generate one or more up-to-date Code Examples for VPlan similar to~\cite{llvmintrvplan,llvmvplanstate,llvmouterloop})
-    $\rightarrow$ Explain different components (e.g.: \texttt{VPBasicBlock}, \texttt{VPRegionBlock}, etc.) and
-    correlation between generated VPlan structure, original scalar code. Show some general optimizations.
-    \item Modelling Cost. Not yet part of VPlan. (related work? already adopted?~\cite{POHL2020102106})
-    \item Optimizing/Transforming VPlans~\cite{llvmvplanupdate}
-  \end{enumerate}
-  \item Phase 3: Execution - Materialize best VPlan
+  \item History
+  \begin{itemize}
+    \item Short Summary how it was before (LoopVectorize)
+    \item Initial refactoring proposition by Intel
+    \item Describe rough final finished VPlan architecture.
+  \end{itemize}
+  \item State of VPlan
+  \begin{itemize}
+    \item Rough algorithm description (interplay of legacy LV, legacy cost model and new VPlan) (with light code references?)
+    \item VPlan structure (HCFG, \texttt{VPRegionBlock}, recipes, etc.) and modelling of IR explained via example similar to~\cite{llvmvplanstate} (What is good example?
+Too simple $\rightarrow$ doesn't cover enough;
+    Too complex $\rightarrow$ doesn't fit into paper.)
+    \item (Outer Loop Vectorization?)
+  \end{itemize}
+  \item Future of VPlan (what is currently being worked on, what is planned next, what has not been planned)
 \end{enumerate}
@@
 \begin{enumerate}
-  \item Other auto-vectorization algorithms / possible improvements~\cite{10.5555/3615924.3615932,4336219,llvmveccontrolflow,saarlandrv,10.1145/3303117.3306172}
-  \\Some of those should already be (partially) implemented in LLVM
-  \item Auto-Vectorization in GCC~\cite{autovecoptgcc, vecoptgcc}
-  \item Comparison with GCC~\cite{gccllvmveccomp,10.1155/2021/3264624}
+  \item Region Vectorizer (RV)~\cite{llvmvplanrv,10.1145/3192366.3192413}
+  \item (Other auto-vectorization algorithms / possible improvements~\cite{10.5555/3615924.3615932,4336219,llvmveccontrolflow,saarlandrv,10.1145/3303117.3306172})
+  \item (Auto-Vectorization in GCC~\cite{autovecoptgcc, vecoptgcc})
+  \item (Comparison with GCC~\cite{gccllvmveccomp,10.1155/2021/3264624})
+  \item (Comparison/Outlook auto-vectorization and explicit vectorization~\cite{10.1155/2021/3264624, 9802745, DBLP:conf/vldb/BensonER23})
   \item (Polyhedral compilation techniques~\cite{llvmpolyhedral})
@@
 \label{sec:summary}
-\begin{enumerate}
-  \item VPlan Summary~\cite{llvmvplan,llvmvplanupdate}
-  \item VPlan Future Roadmap~\cite{llvmvplan,llvmvplanupdate} (e.g.: combining outer-/inner-loop paths, full function vectorization, etc.)
-  \item (Comparison/Outlook auto-vectorization and explicit vectorization~\cite{10.1155/2021/3264624, 9802745, DBLP:conf/vldb/BensonER23})
-\end{enumerate}
+VPlan Summary
 

### 2024-11-23 paper/paper.tex
paper

@@
 \begin{abstract}
-Lorem ipsum
+  Modern SIMD processors provide various vector registers and ISAs for programmers to utilize. 
+  However, manually vectorizing code can be time-consuming, so compilers look to auto-vectorize 
+  code for better performance. LLVM, as a widely used compiler toolchain, could only auto-vectorize 
+  basic scalar code. Intel proposed a comprehensive refactoring of the underlying system to enable 
+  LLVM's auto-vectorization to support more complex optimizations. The suggested Vectorization Plan 
+  (VPlan) architecture would be more modular and scalable. Since then, significant efforts have been 
+  made to implement this new system. Today, VPlan is already actively involved in vectorizing 
+  innermost loops. Other areas, such as outer loop vectorization, are also in development.
 \end{abstract}
@@
 
-While explicit data dependencies are easier to see, even a simple loop, as in 
-Figure~\ref{fig:inner-loop-vec}, could carry an implicit dependency in line 3, if
+While explicit data dependencies are easier to see, a simple loop, as in 
+Figure~\ref{fig:inner-loop-vec}, could carry an implicit dependency in line 3 if
 the \texttt{x} and \texttt{y} point to overlapping memory regions (pointer aliasing).
@@
 \subsection{Vectorization Costs}
-While it may not be legal to vectorize a given control flow, the vectorized code sometimes 
-performs worse than its scalar equivalent.
+While vectorizing a given control flow may not be legal, the vectorized code sometimes performs 
+worse than its scalar equivalent.
 
@@
 and SLPVectorize. These vectorizers are a part of LLVM's middle-end and run after module 
-simplification has taken place. As VPlan has not yet been integrated into SLP-Vectorization 
-we will only focus on explaining the original LoopVectorize structure.
+simplification has taken place. As VPlan still needs to be integrated into SLP-Vectorization, we 
+will only focus on explaining the original LoopVectorize structure.
 
 LoopVectorize was designed to focus only on vectorization of innermost loops~\cite{llvmintrvplan}. 
-When starting with a initial scalable loop in IR it would first run a legality phase to check 
-whetever and how it is legal to vectorize the loop. This phase would produce artifacts/maps like if 
-and what runtime aliasing checks would be required. Then the cost model ran next, creating more 
+When starting with an initial scalable loop in IR, it would first run a legality phase to check 
+whether  and how it is legal to vectorize the loop. This phase would produce artifacts/maps, such as if 
+and what runtime aliasing checks would be required. Then, the cost model ran next, creating more 
 artifacts based on the scalar IR and legality artifacts. These cost model artifacts would 
-associate different vectorization factors with a cost. Lastly the transformation phase would 
-vectorize the underlying IR in one go based on all the produced artifacts.
+associate different vectorization factors with a cost. Lastly, the transformation phase would vectorize 
+the underlying IR in one go based on all the produced artifacts.
 
-However, this rigid flow of LoopVectorize was a problem. Extending LV to include outer loop 
-and nested loop vectorizations would have been very challenging, as LV was designed to flatten a
-loops control flow into a single basic block. Producing and considering more and 
-more artifacts when trying to transform the scalar loop into a vectorized one all at once did also 
+However, this rigid flow of LoopVectorize presented a challenge. Extending LV to include outer loop 
+and nested loop vectorizations would have been difficult, as LV was designed to flatten a
+loop's control flow into a single basic block. Producing and considering more and 
+more artifacts when trying to transform the scalar loop into a vectorized one all at once also did 
 not scale well.
 
-As a potential solution Intel proposed a long-term effort to refactor LLVM's whole auto-vectorization
+As a potential solution, Intel proposed a long-term effort to refactor LLVM's whole auto-vectorization
 using a system called Vectorization Plan (VPlan)~\cite{llvmextloopvec}.
@@
 The goal was to unite any auto-vectorization systems (SLP, Inner-/Outer-Loop) into a single flexible
-system~\cite{llvmintrvplan,llvmvplanstate}, as seen in Figure~\ref{fig:vplan-future}. This system would use VPlans to represent different 
-vectorization approaches, abstracted away from the underlying IR. Each VPlan would be able to 
-calculate its cost and transform itself to modify the final IR. At first the Legality 
-phase (yellow) would check if the scalar code would be legal to vectorize at all and if so initial 
-VPlans would be constructed. In the following Planning phase (magenta) these initial VPlans would 
-be optimized using various VPlan-to-VPlan transformations~\cite{llvmouterloopstatus}. 
-Those transformations would be supervised 
-by the VPlan's cost model. Finally, the best VPlan would be chosen and materialized/executed, 
-modifying the underlying IR. The best VPlan might also be to not vectorize at all (VF $=1$).
+system~\cite{llvmintrvplan,llvmvplanstate}, as seen in Figure~\ref{fig:vplan-future}. This system would 
+use VPlans to represent different 
+vectorization approaches abstracted away from the underlying IR. Each VPlan would 
+calculate its cost and transform itself to modify the final IR. At first, the Legality phase (yellow) 
+would check if the scalar code would be legal to vectorize at all, and if so, initial VPlans would 
+be constructed. In the following Planning phase (magenta), these initial VPlans would be optimized 
+using various VPlan-to-VPlan transformations~\cite{llvmouterloopstatus}. 
+The VPlan's cost model would supervise those transformations. Finally, the best VPlan would be chosen 
+and materialized/executed, modifying the underlying IR.
 
@@
 
-In its current state VPlan is deeply integrated into LV.
+In its current state, VPlan is deeply integrated into LV.
 When auto-vectorizing, LV starts by checking all the different loops in the IR.
@@
 Several factors might hinder vectorization for a given loop: 
-(1) If the loop was already vectorized before, it won't be vectorized again. 
-(2) The Legality phase checks if an how it is legal to vectorize the scalar
+(1) If the loop was already vectorized before, it will not be vectorized again. 
+(2) The Legality phase checks if and how it is legal to vectorize the scalar
 loop. (3) The loop's trip count might be too small to make vectorization profitable. (4) The cost 
-of runtime checks for e.g. pointer aliasing can be too high.
+of runtime checks, such as pointer aliasing, can be too high.
 
-The initial VPlans are constructed based on the legality checks and the cost decisions the legacy
-LV cost model made referencing the scalar IR (\texttt{LoopVectorizationPlanner::\allowbreak plan}).
-After initial VPlan construction various VPlan-to-VPlan transformations are applied to optimize
-it (\texttt{LoopVectorizationPlanner::\allowbreak buildVPlansWithVPRecipes}).
+The initial VPlans are constructed based on the legality checks and the legacy LV cost model's 
+cost decisions referencing the scalar IR (\texttt{LoopVectorizationPlanner::\allowbreak plan}). 
+After initial VPlan construction, various VPlan-to-VPlan 
+transformations are applied to optimize it 
+(\texttt{LoopVectorizationPlanner::\allowbreak buildVPlansWithVPRecipes}).
 
-Then the optimal Vectorization Factor (VF) is calculated based on the set of VPlans and their
-representive cost (\texttt{LoopVectorizationPlanner::\allowbreak cost}), 
-which calculation still involves the legacy LV cost model. 
-The best VPlan is picked based on the optimal VF. A VF of 1 means it is not
-profitable to vectorize the loop, however, interleaving/unrolling can still end up modifying the
-underlying IR.
-Afterwards, the Interleave Count (IC) is determined 
+Then, the optimal vectorization factor (VF) is calculated based on the set of VPlans and their 
+representative cost (\texttt{LoopVectorizationPlanner::\allowbreak cost}), the calculation of 
+which still involves the legacy LV cost model.
+The best VPlan is picked based on the optimal VF. A VF of 1 means it is not profitable to vectorize 
+the loop; however, interleaving/unrolling can still modify the underlying IR. Afterwards, the 
+Interleave Count (IC) is determined 
 (\texttt{LoopVectorizationCostModel::\allowbreak selectInterleaveCount}).
 
-The best VPlan is then executed (\texttt{LoopVectorizationPlanner::\allowbreak executePlan}). In case of
-vectorization (VF $>1$), LV will also directly try to vectorize the epilogue loop.
+The best VPlan is then executed (\texttt{LoopVectorizationPlanner::\allowbreak executePlan}). In the 
+case of vectorization (VF $>1$), LV will also directly try to vectorize the epilogue loop.
 
@@
 \texttt{if.then}, and \texttt{for.inc}. Figure~\ref{fig:inner-loop-vplan-hcfg-body} shows the equivalent
-hierarchical control-flow-graph (CFG) of a VPlan modelling a vectorization approach for this 
-scalar loop. Other basic blocks like the scalar epilogue loop and it's predication check (is the final
+hierarchical control-flow-graph (CFG) of a VPlan modeling a vectorization approach for this 
+scalar loop. Other basic blocks like the scalar epilogue loop and its predication check (is the final
 vector trip count equal to the original trip count) are omitted. See Figure~\ref{fig:inner-loop-vplan-hcfg}
@@
 \label{sec:relatedwork}
-Lorem ipsum
+Besides the main llvm-project's upstream there are many different projects making use of or extending
+LLVM's toolchain.
 

### 2024-11-23 paper/paper.tex
paper

@@
   \includegraphics[width=0.475\textwidth]{images/inner-loop-scalar-loop-ir-color.png}
-  \caption{The scalar IR for the main loop part of the conditional loop depicted in
+  \caption{The scalar IR for the main loop section of the conditional loop depicted in
   Figure~\ref{fig:inner-loop-vec}. Note that \texttt{getelementptr inbounds}
@@
   \includegraphics[width=0.475\textwidth]{images/inner-loop-vplan-hcfg-loop-body-color.png}
-  \caption{The vectorized loop part of a hierarchical CFG of a VPlan modelling the conditional loop
+  \caption{The vectorized loop section of the hierarchical CFG of a VPlan modeling the conditional loop
   depicted in Figure~\ref{fig:inner-loop-vec}. Some predefined values are
@@
 
-The different colored parts in the scalar IR can be found with the same color in the modeled VPlan.
-The only completely new part that VPlan models in this example is the handling of the scalar epilogue
-loop (purple).
+The different colored parts in the scalar IR map to the modeled VPlan region with the same color.
+The only new part that VPlan models in this example is handling the scalar epilogue loop (purple).
 
@@
 are either \texttt{VPBasicBlocks}, containing a sequence of one or more \texttt{VPRecipes}, or
-\texttt{VPRegionBlocks}, modelling a single-entry-single-exit subgraph.
-In our example the \texttt{vector.body} is a basic block and \texttt{vector loop} is a region block.
+\texttt{VPRegionBlocks}, modeling a single-entry-single-exit subgraph.
+In the example, \texttt{vector loop} is a region block. All other blocks are basic blocks.
 
-\texttt{VPRegionBlocks} can indicate that their contents are to be replicated a certain number of times,
-to model more complex control flow for different VF, UF combinations more efficiently in a single VPlan. 
-In our example \texttt{vector loop} is not to be replicated (\texttt{<x1>}).
+\texttt{VPRegionBlocks} can indicate that their contents are to be replicated a certain number of times
+to model more complex control flow for different VF-UF combinations more efficiently in a single VPlan. 
+In Figure~\ref{fig:inner-loop-vplan-hcfg-body}, \texttt{vector loop} is not to be replicated (\texttt{<x1>}).
 
-Inside our two \texttt{VPBasicBlocks} we have a list of recipes (lines of instructions). A recipe
-models one or more instructions in the executed IR and can be based on one or more instructions.
-from the input IR (ingredients). Note that a recipe might also model a instruction with no reference to any input
-IR (in our example this would be the inversion of our vector mask). There are different types of 
+Within our two \texttt{VPBasicBlocks} are lists of recipes (lines of instructions). A recipe
+models one or more instructions in the executed IR and can be based on one or more instructions
+from the input IR (ingredients). Note that a recipe might also model an instruction with no reference to any input
+IR (in our example, this would be the inversion of our vector mask). There are different types of 
 recipes. \texttt{VPReplicateRecipe} will either \texttt{CLONE} exactly one instruction from the input IR
@@
 \texttt{VPWidenRecipes} exist to transform/\texttt{WIDEN} any input IR instruction into a vectorized 
-version of it. The last remaining recipe type in our example are \texttt{VPInstructions}. They
-\texttt{EMIT} new instructions into the output IR that have no representative in the input IR (no
+version of it. The last remaining recipe type in our example is \texttt{VPInstruction}. This recipe
+\texttt{EMIT}s new instructions into the output IR that have no representative in the input IR (no
 ingredients).
@@
 Furthermore, VPlan recipes can also model definitions and usages of different values. In our example,
-the comparison mask \texttt{vp<\%6>} is a \texttt{VPValue} that is used by three different 
-\texttt{VPUsers}.
+the comparison mask \texttt{vp<\%6>} is a \texttt{VPValue} used by three \texttt{VPUsers}.
 
 \subsection{Future of VPlan}
-Many of the VPlan internals and systems surrounding it are still in active development. To stay
-up-to-date with the latest developments regarding auto-vectorization in LLVM one can follow the 
+Many of the VPlan internals and surrounding systems are still in active development. To stay up-to-date 
+with the latest developments regarding auto-vectorization in LLVM, one can follow the 
 Community Forum\footnote{\url{https://discourse.llvm.org/c/ir-optimizations/loop-optimizations/62}} or
@@
 \subsubsection{VPlan and Integration with LV}
-The main effort of setting up the VPlan structure and integrating it first into LV has seen
-tremendous effort over the last years. Florian Hahn presented future directions for 
-VPlan~\cite{llvmvplanupdate}, which mostly evolve around transforming more LV-based decisions into
-explicit VPlan-to-VPlan transformations. Additionally, a lot of cost model related decisions 
-still end up being influenced by the old legacy LV cost model, however, transition to a VPlan-based 
-cost model has already started.
+Setting up and integrating the VPlan structure into LV has seen tremendous effort over the last 
+few years. Florian Hahn presented future directions for 
+VPlan~\cite{llvmvplanupdate}, which mostly revolve around transforming more LV-based decisions into
+explicit VPlan-to-VPlan transformations. Additionally, the old legacy LV cost model still influences 
+many cost model-related decisions. However, the transition to a VPlan-based cost model has already 
+started.
 
 \subsubsection{Outer Loop Vectorization}
-Development around outer loop vectorization has slowed down a bit as focus shifted more onto VPlan
-internals and integration into LV. The inner loop path and outer loop path are currently still
-seperated by the experimental \texttt{-enable-\allowbreak vplan-\allowbreak native-\allowbreak path} 
-flag. Further developing the 
-outer loop path and eventually merging it with the innermost loop path is planned.
+Development around outer loop vectorization has slowed as the focus shifted to VPlan internals 
+and integration into LV. The inner and outer loop paths are still separated, and the 
+experimental \texttt{-enable-\allowbreak vplan-\allowbreak native-\allowbreak path} flag must be 
+set to activate any outer 
+loop vectorization. Further development of the outer loop path and eventually merging it 
+with the innermost loop path is planned.
 
 \subsubsection{Merging with SLP Vectorize}
-Merging the existing SLP Vectorizer with the current VPlan infrastructure is in the early phase
-of planning.
+Merging the existing SLP Vectorizer with the current VPlan infrastructure is in the early planning phase.
 
 \subsubsection{Function Vectorization}
-As far as we now there is currently no active development for whole-function vectorization in the
-llvm-project's upstream yet.
+There is currently no active development for whole-function vectorization in the
+llvm-project's upstream.
 
@@
 \label{sec:summary}
-In conclusion, VPlan has been in active development for several years now. When compiling with
-LLVM VPlan is used to model initial vectorization approaches, keep track of different
-optimization decisions and finally execute the best VPlan to transform IR. 
-However, there are still some VPlan-to-VPlan transformations and a VPlan-based cost model missing, 
-before the legacy LV code can be completely discarded.
+In conclusion, VPlan has been in active development for several years now. When compiling with LLVM, 
+VPlans model initial vectorization approaches, keep track of different optimization decisions, 
+and finally, the best VPlan executes to transform the underlying IR. 
+However, some VPlan-to-VPlan transformations and a VPlan-based cost model must be included before the 
+legacy LV code can be discarded entirely.
 
-Outer loop vectorization support via VPlan is still experimental, seperated from the inner loop
+Outer loop vectorization support via VPlan is still experimental, separated from the inner loop
 LV integration, and runs on a divergent experimental path (\texttt{-enable-\allowbreak vplan-\allowbreak native-\allowbreak path}).
-In the near future development on outer loop vectorization will continue and merging it with LV's
-VPlan path is planned.
+In the near future, development on outer loop vectorization will continue, and merging 
+it with LV's VPlan path is planned.
 
-Active development of other approaches such as SLP and whole-function vectorization based on the 
-VPlan architecture has not been started yet.
+Active development of other approaches, such as SLP and whole-function vectorization based on the VPlan 
+architecture, has yet to be started.
 

### 2024-11-23 paper/paper.tex
paper

@@
 VPlan~\cite{llvmvplanupdate}, which mostly revolve around transforming more LV-based decisions into
-explicit VPlan-to-VPlan transformations. Additionally, the old legacy LV cost model still influences 
+explicit VPlan-to-VPlan transformations. Additionally, the legacy LV cost model still influences 
 many cost model-related decisions. However, the transition to a VPlan-based cost model has already 
@@
 \label{sec:relatedwork}
-Besides the main llvm-project's upstream there are many different projects making use of or extending
-LLVM's toolchain.
+Next to the main llvm-projects upstream, many projects are using or extending the LLVM toolchain.
+
+One such project that also tries to enable auto-vectorization support for inner-, outer-loop, and 
+whole-function vectorization is the region vectorizer from the University Saarland~\cite{rv}.
+This region-vectorizer implements partial control-flow linearization~\cite{10.1145/3192366.3192413}
+and a divergence analysis that can handle unstructured CFGs. There has been an effort to partially
+implement this control-flow linearization and divergence analysis into LLVM's 
+upstream~\cite{rvproposal,rvproposaldep}.
 

### 2024-12-04 paper/paper.tex
paper

@@
 \begin{document}
-\title{Utilizing Parallel Workers: \\LLVM's Vectorization Plan}
+\title[State of LLVM's Vectorization Plan]{Utilizing Parallel Workers: LLVM's Vectorization Plan}
 \author{Jonas Fritsch}
@@
   innermost loops. Other areas, such as outer loop vectorization, are also in development.
+  In this paper we give an introduction into VPlan and summarize its latest development results.
 \end{abstract}
@@
 in a single flexible system capable of optimizing SLPs, inner, and nested loops with complex 
-control flows. The auto-vectorization pass would create and transform multiple different 
-VPlans, each modeling a different vectorization approach, abstracted away from the underlying LLVM IR.
-Those VPlans are compared against each other based on a cost model, and finally, the best one, which
-might also not vectorize at all, is materialized into the LLVM IR.
+control flow.
+
+The auto-vectorization pass would create multiple different VPlans, based on initial legality 
+checks. Each VPlan represents a different vectorization approaches, by modeling the
+underlying scalar control flow with its own abstracted vectorized version. Different VPlan-to-VPlan
+transformations can then optimize these abstracted control flows based on a cost model.
+After that, the best VPlan, and with it the best vectorization approach, is chosen by the cost model.
+This VPlan is then executed, by going over its own vectorized control flow representation and
+materializing it back into the underlying LLVM IR.
 
@@
 
+The rest of this paper is organized as follows. We first explain different vectorization
+strategies and vectorization constraints in general and describe the original auto-vectorization
+approach in LLVM. Section~\ref{sec:vplan} then introduces the new VPlan approach, focusing on
+its most important individual components and gives an outlook on future developments. In 
+Section~\ref{sec:relatedwork} we highlight related work to LLVM's auto-vectorization. Concluding,
+we summarize VPlan, its current state and potential future.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \label{sec:background}
+
+\subsection{Vectorization Strategies}
+\label{back:vec-strat}
 In general, vectorization techniques can be separated into different categories:
 
-\subsection{Inner Loop Vectorization}
-Innermost loops are loops whose body's control flow does not contain any more loops, bound or unbound.
+\paragraph{Inner Loop Vectorization}
+Innermost loops are loops whose body's control flow does not contain any more loops.
 Vectorization of these loops is more straightforward as the only option to vectorize is over 
@@
 
-\subsection{Super-Word Parallelism (SLP) Vectorization}
-SLP Vectorization vectorizes similar independent instructions. A compiler might have, for example, 
-fully unrolled a smaller loop in a previous code simplification pass. The resulting similar scalar 
+\paragraph{Super-Word Parallelism (SLP) Vectorization}
+SLP Vectorization vectorizes similar independent sequential instructions. This can often be seen 
+after a loop has been partially or fully unrolled. The resulting similar scalar 
 instructions might then be vectorized later by an SLP vectorizer.
 
-\subsection{Outer Loop Vectorizaton}
-Outer loop vectorization focuses on control flow with nested loops. Such loops often present
+\paragraph{Outer Loop Vectorizaton}
+Outer loop vectorization focuses on control flow with nested loops. Such control flow often presents
 more challenges for auto-vectorization, as one could vectorize over the outer loop,
 inner loop, or a mix of both, representing a hybrid/multi-dimensional vectorization approach.
-Carrying out and comparing costs of such different vectorization approaches can be complex 
-and require a flexible system like, for example, VPlans.
+Finding the best legal vectorization can therefor be more difficult.
 
@@
   \caption{A code snippet of a nested loop where it would be beneficial to vectorize over the 
-  outer loop. The inner loop has a trip count that is too low (6) to vectorize. Additionally,
+  outer loop. The inner loop has a trip count (6) that is too low to vectorize. Additionally,
   the memory access pattern of \texttt{y} would be consecutive compared to scattered when
@@
 
-\subsection{Function Vectorization}
+\paragraph{Function Vectorization}
 Function vectorization refers to the vectorization of entire functions. A loop might contain a
@@
 this call could be replaced with a call to a newly constructed function with the same semantic 
-control flow but operating on multiple values at once through vector registers instead.
-
-\subsection{Vectorization Similarities}
-All of the above-mentioned vectorization techniques can be brought into relation with each other. 
-For example, function vectorization can be transformed into loop vectorization by creating a loop 
-around the function body and vectorizing it and vice versa. The same applies for SLP Vectorization 
-and loop vectorization when unrolling a loop.
+control flow but operating on multiple values at once through vector registers.
 
 \subsection{Vectorization Legality}
+\label{sec:vec-legal}
 Not all loops can be vectorized. There are various factors that can ultimately hinder any
@@
 
-Not all data dependencies directly hinder any vectorization. One exception can 
+Not all data dependencies directly hinder any vectorization~\cite{datadepvec}. An exception can, for example, 
 sometimes be made for reduction idioms, as seen in the inner loop in 
@@
 
-Additionally, the control flow should be primarily straight-line code. While most if-else 
-constructs can be flattened to always execute both branches along with masking, the same is not 
-valid for switch-case statements or too complex control flows.
+Uncountable loops can typically also not be vectorized, due to their unknown number of iterations.
+
+Finally some instructions like integer division do inherently not have a vectorized instruction equivalent.
 
@@
 due to many factors: (1) The transformed control flow often includes more instructions.
-(2) As the loop trip count might not be a perfect multiple of the vectorization factor, a scalar
+(2) As the loop trip count might not be a perfect multiple of the Vectorization Factor, a scalar
 epilogue/remainder of the loop might be used to handle the tail iterations. (3) Various runtime
@@
 to overlapping memory regions, in which case the scalar version of the loop is executed.
+
+\subsection{Auto-Vectorization in LLVM}
+LLVM implements two different vectorizers for auto-vectorization, LoopVectorize (LV)
+and SLPVectorize~\cite{llvmvec,llvmhistorystate}. Auto-Vectorization is part of LLVM's 
+middle-end and runs after module simplification has taken place.
+In general, these vectorizers take in scalar LLVM IR and transform it into optimized vectorized 
+LLVM IR.
+
+SLPVectorize is a SLP vectorizer working on straight-line code (non-loops). It works by analyzing
+the code bottom-up, keeping track of dependencies and grouping similar independent scalar instructions
+together if possible. Based on a cost model SLPVectorize then decides to vectorize these groups.
+
+LoopVectorize was designed to focus only on vectorization of innermost loops~\cite{llvmintrvplan}. 
+When starting with an initial scalable loop in IR, it would first run a legality phase to check 
+whether and how it is legal to vectorize the loop. This phase would produce different artifacts, 
+such as if 
+and what runtime aliasing checks would be required. Then, the cost model ran next, creating more 
+cost-based artifacts based on the underlying scalar IR and the legality artifacts. 
+Lastly, the transformation phase would vectorize 
+the IR in one go based on all the previously produced artifacts.
+
+However, this rigid flow of LoopVectorize presented a challenge. Extending LV to include outer loop 
+and nested loop vectorizations would have been difficult, as LV was designed to flatten a
+loop's control flow into a single basic block. Producing and considering more and 
+more artifacts when trying to transform the scalar loop into a vectorized one all at once also did 
+not scale well.
+
+As a potential solution, Intel proposed a long-term effort to refactor LLVM's whole auto-vectorization
+using a more modular and flexible system called Vectorization Plan (VPlan)~\cite{llvmextloopvec}.
 
@@
 
-\subsection{History}
-LLVM implements two different vectorizers for auto-vectorization, LoopVectorize (LV)
-and SLPVectorize. These vectorizers are a part of LLVM's middle-end and run after module 
-simplification has taken place. As VPlan still needs to be integrated into SLP-Vectorization, we 
-will only focus on explaining the original LoopVectorize structure.
+The goal of Intels initial VPlan proposal was to unite any 
+auto-vectorization systems (SLP, Inner-/Outer-Loop, Whole-Function) into a single flexible
+system~\cite{llvmextloopvec,llvmintrvplan,llvmvplanstate}. 
+This system would use VPlans to represent different 
+vectorization approaches, abstracting away from the underlying scalar loop IR by using its own 
+hierarchical control-flow-graph (H-CFG). This allows for iteratively exploring different 
+optimization decisions by performing transformations on this CFG without modifying any LLVM IR.
+Each VPlan would be able to calculate its cost and execute itself, materializing the vectorization 
+approach back into LLVM IR.
 
-LoopVectorize was designed to focus only on vectorization of innermost loops~\cite{llvmintrvplan}. 
-When starting with an initial scalable loop in IR, it would first run a legality phase to check 
-whether  and how it is legal to vectorize the loop. This phase would produce artifacts/maps, such as if 
-and what runtime aliasing checks would be required. Then, the cost model ran next, creating more 
-artifacts based on the scalar IR and legality artifacts. These cost model artifacts would 
-associate different vectorization factors with a cost. Lastly, the transformation phase would vectorize 
-the underlying IR in one go based on all the produced artifacts.
-
-However, this rigid flow of LoopVectorize presented a challenge. Extending LV to include outer loop 
-and nested loop vectorizations would have been difficult, as LV was designed to flatten a
-loop's control flow into a single basic block. Producing and considering more and 
-more artifacts when trying to transform the scalar loop into a vectorized one all at once also did 
-not scale well.
-
-As a potential solution, Intel proposed a long-term effort to refactor LLVM's whole auto-vectorization
-using a system called Vectorization Plan (VPlan)~\cite{llvmextloopvec}.
-
-The goal was to unite any auto-vectorization systems (SLP, Inner-/Outer-Loop) into a single flexible
-system~\cite{llvmintrvplan,llvmvplanstate}, as seen in Figure~\ref{fig:vplan-future}. This system would 
-use VPlans to represent different 
-vectorization approaches abstracted away from the underlying IR. Each VPlan would 
-calculate its cost and transform itself to modify the final IR. At first, the Legality phase (yellow) 
-would check if the scalar code would be legal to vectorize at all, and if so, initial VPlans would 
-be constructed. In the following Planning phase (magenta), these initial VPlans would be optimized 
-using various VPlan-to-VPlan transformations~\cite{llvmouterloopstatus}. 
-The VPlan's cost model would supervise those transformations. Finally, the best VPlan would be chosen 
-and materialized/executed, modifying the underlying IR.
+This planned systems is shown in Figure~\ref{fig:vplan-future}. At first, the Legality 
+Phase would check if the scalar code would be legal to vectorize at all, and if so, 
+initial VPlans would be constructed. In the following Planning Phase, these initial 
+VPlans would be optimized using various VPlan-to-VPlan transformations~\cite{llvmouterloopstatus}. 
+The VPlan's cost model would supervise those transformations. Finally, the best VPlan 
+would be chosen and executed, modifying the underlying IR.
 
@@
   \includegraphics[width=0.475\textwidth]{images/vplan-future.png}
-  \caption{The planned future architecture for VPlan. The VPlan optimization would
-  work as iteratively applying VPlan-to-VPlan transformations.}
+  \caption{The planned future architecture for auto-vectorization in LLVM. The 
+  optimization step in phase 2 would iteratively apply VPlan-to-VPlan transformations.}
   \label{fig:vplan-future}
@@
 
-\subsection{State of VPlan (LLVM 19.1.4)}
-\emergencystretch 6em
+\subsection{VPlan Structure}
+Before explaining how far this system is already implemented in LLVM we will go over
+the different elements of a \texttt{VPlan} in more detail.
 
-In its current state, VPlan is deeply integrated into LV.
-When auto-vectorizing, LV starts by checking all the different loops in the IR.
-If it finds one, it will try to vectorize it (\texttt{LoopVectorizePass::\allowbreak processLoop}).
-Several factors might hinder vectorization for a given loop: 
-(1) If the loop was already vectorized before, it will not be vectorized again. 
-(2) The Legality phase checks if and how it is legal to vectorize the scalar
-loop. (3) The loop's trip count might be too small to make vectorization profitable. (4) The cost 
-of runtime checks, such as pointer aliasing, can be too high.
+The purpose of a single VPlan object is to model a candidate for vectorization. 
+Such a single candidate can potentially represent multiple different vectorized loops in the end. 
+For example, one VPlan might model a general vectorization approach for multiple different 
+Vectorization Factors (VFs) and multiple different Unroll Factors (UFs) if applicable.
 
-The initial VPlans are constructed based on the legality checks and the legacy LV cost model's 
-cost decisions referencing the scalar IR (\texttt{LoopVectorizationPlanner::\allowbreak plan}). 
-After initial VPlan construction, various VPlan-to-VPlan 
-transformations are applied to optimize it 
-(\texttt{LoopVectorizationPlanner::\allowbreak buildVPlansWithVPRecipes}).
+\paragraph{H-CFG}
+This is achieved by introducing a new layer of abstraction from the underlying loop's scalar IR.
+Each VPlan stores its own Hierarchical-CFG (H-CFG) modeling the vectorization candidate. This 
+control-flow graph and its components are the main element of a single VPlan.
+The nodes of this H-CFG are one of three types:
 
-Then, the optimal vectorization factor (VF) is calculated based on the set of VPlans and their 
-representative cost (\texttt{LoopVectorizationPlanner::\allowbreak cost}), the calculation of 
-which still involves the legacy LV cost model.
-The best VPlan is picked based on the optimal VF. A VF of 1 means it is not profitable to vectorize 
-the loop; however, interleaving/unrolling can still modify the underlying IR. Afterwards, the 
-Interleave Count (IC) is determined 
-(\texttt{LoopVectorizationCostModel::\allowbreak selectInterleaveCount}).
+\begin{itemize}
+  \item \texttt{VPBasicBlock} stores a list of \texttt{VPRecipes}, similar to 
+  the LLVM IR's \texttt{BasicBlock} holding a list of IR Instructions.
+  \item \texttt{VPRegionBlock} represents a subgraph with a single VP entry node and a 
+  single VP exit node. A \texttt{VPRegionBlock} can be marked to replicate its subgraph
+  \texttt{VF * UF} times when executing. This is useful for modeling scalarized or predicated
+  instructions within the final vectorized loop body.
+  \item \texttt{VPIRBasicBlock} is a special \texttt{VPBasicBlock} that wraps an underlying
+  IR \texttt{BasicBlock}. These nodes are useful for modeling the scalar parts around the 
+  final vectorized loops such as the loop preheader and the epilogue loop.
+\end{itemize}
 
-The best VPlan is then executed (\texttt{LoopVectorizationPlanner::\allowbreak executePlan}). In the 
-case of vectorization (VF $>1$), LV will also directly try to vectorize the epilogue loop.
-
-\subsubsection{VPlan Structure}
-To better understand the structure of a VPlan, we will use the
-simple conditional loop from Figure~\ref{fig:inner-loop-vec} as an example.
-
-Figure~\ref{fig:inner-loop-scalar-ir} shows the scalar IR of the main loop basic blocks \texttt{for.body},
-\texttt{if.then}, and \texttt{for.inc}. Figure~\ref{fig:inner-loop-vplan-hcfg-body} shows the equivalent
-hierarchical control-flow-graph (CFG) of a VPlan modeling a vectorization approach for this 
-scalar loop. Other basic blocks like the scalar epilogue loop and its predication check (is the final
-vector trip count equal to the original trip count) are omitted. See Figure~\ref{fig:inner-loop-vplan-hcfg}
-in Appendix~\ref{appendix-figures} for the fully modeled HCFG.
-
-%TODO: Combine inner-loop-scalar-ir figure and inner-loop-vplan-hcfg into a single
-%       double column figure.
+Figure~\ref{fig:inner-loop-scalar-ir} shows the loop part of the scalar IR for the code example
+of Figure~\ref{fig:inner-loop-vec}. Comparatively, Figure~\ref{fig:inner-loop-vplan-hcfg-body}
+shows the loop part of the H-CFG of a VPlan that models the vectorization approach described
+in Figure~\ref{fig:inner-loop-vec}.
 
@@
   \includegraphics[width=0.475\textwidth]{images/inner-loop-vplan-hcfg-loop-body-color.png}
-  \caption{The vectorized loop section of the hierarchical CFG of a VPlan modeling the conditional loop
-  depicted in Figure~\ref{fig:inner-loop-vec}. Some predefined values are
+  \caption{The vectorized loop section of the hierarchical CFG of a VPlan modeling the 
+  conditional loop depicted in Figure~\ref{fig:inner-loop-vec}. 
+  Surrounding \texttt{VPIRBasicBlock}-nodes are omitted.
+  Some predefined values are
   \texttt{vp<\%0>} = VF * UF; \texttt{vp<\%1>} = vector-trip-count; \texttt{vp<\%2>} = original trip-count.}
@@
 
-The different colored parts in the scalar IR map to the modeled VPlan region with the same color.
-The only new part that VPlan models in this example is handling the scalar epilogue loop (purple).
+\paragraph{Recipes}
 
-The hierarchical CFG of a VPlan helps abstract it away from the underlying IR. The nodes of this CFG
-are either \texttt{VPBasicBlocks}, containing a sequence of one or more \texttt{VPRecipes}, or
-\texttt{VPRegionBlocks}, modeling a single-entry-single-exit subgraph.
-In the example, \texttt{vector loop} is a region block. All other blocks are basic blocks.
+\texttt{VPRecipes} can be seen as instructions of an IR specialized for auto-vectorization. 
+A single \texttt{VPRecipe} inside a VPlan is based on zero or more LLVM IR instructions of the 
+underlying scalar IR and will materialize into one or more LLVM IR instructions when executed.
 
-\texttt{VPRegionBlocks} can indicate that their contents are to be replicated a certain number of times
-to model more complex control flow for different VF-UF combinations more efficiently in a single VPlan. 
-In Figure~\ref{fig:inner-loop-vplan-hcfg-body}, \texttt{vector loop} is not to be replicated (\texttt{<x1>}).
+There are many different types of \texttt{VPRecipes}, most notably:
+\begin{itemize}
+  \item \texttt{VPReplicateRecipe} will replicate one IR instruction either once (\texttt{CLONE})
+  or multiple times (\texttt{REPLICATE}) into the vectorized IR when executed.
+  \item \texttt{VPWidenRecipe} will \texttt{WIDEN} one IR instruction into a
+  vector-equivalent when executed.
+  \item \texttt{VPInstruction} has no underlying IR instruction it is based on and is used 
+  to model newly needed instructions for the vectorized IR. When executed it will 
+  \texttt{EMIT} one or more new IR instructions.
+\end{itemize}
 
-Within our two \texttt{VPBasicBlocks} are lists of recipes (lines of instructions). A recipe
-models one or more instructions in the executed IR and can be based on one or more instructions
-from the input IR (ingredients). Note that a recipe might also model an instruction with no reference to any input
-IR (in our example, this would be the inversion of our vector mask). There are different types of 
-recipes. \texttt{VPReplicateRecipe} will either \texttt{CLONE} exactly one instruction from the input IR
-into exactly one instruction in the output IR, or \texttt{REPLICATE} an instruction a certain number
-of times into output IR (typically found in replicator \texttt{VPRegionBlocks}). Different 
-\texttt{VPWidenRecipes} exist to transform/\texttt{WIDEN} any input IR instruction into a vectorized 
-version of it. The last remaining recipe type in our example is \texttt{VPInstruction}. This recipe
-\texttt{EMIT}s new instructions into the output IR that have no representative in the input IR (no
-ingredients).
+The \texttt{VPRecipes} in Figure~\ref{fig:inner-loop-vplan-hcfg-body} have a \texttt{PREFIX}
+infront of them indicating their type. Additionally instructions/recipes of the same color 
+between Figure~\ref{fig:inner-loop-scalar-ir} and Figure~\ref{fig:inner-loop-vplan-hcfg-body}
+are related. The only recipes in our example VPlan CFG that do not semantically model any 
+instructions found in the scalar IR are from the \texttt{middle.block} (purple). This block 
+is used to decide if the scalar epilogue loop is to be executed following the vectorized loop.
 
-Furthermore, VPlan recipes can also model definitions and usages of different values. In our example,
-the comparison mask \texttt{vp<\%6>} is a \texttt{VPValue} used by three \texttt{VPUsers}.
+\paragraph{Transformations}
+Another main benefit of the modular VPlan system is that most vector-optimization cases can 
+be directly implemented as standalone VPlan-to-VPlan transformations. A VPlan-to-VPlan 
+transformation is a static method that either transforms a given VPlan or creates a new 
+VPlan based on a given VPlan. The analysis for these transformations can also be done 
+directly on the VPlan via its \texttt{VPRecipes}.
+
+\subsection{State of VPlan (LLVM 19.1.4)}
+\emergencystretch 6em
+In its current state, VPlan is already deeply integrated into LoopVectorize (LV) and is
+used to model, optimize and execute vectorization of innermost loops~\cite{llvmvplan,llvmintrvplan,llvmvplanupdate}. 
+While outer loop vectorization via VPlan has already seen quite some development, any 
+outer loop vectorization features are still experimental and need to be explicitly enabled 
+by passing the \texttt{-enable-\allowbreak vplan-\allowbreak native-\allowbreak path} 
+feature flag~\cite{llvmouterloop,llvmouterloopstatus}.
+
+To better understand how VPlan is already being used in LLVM, we will now follow along the 
+LoopVectorize pass and summarize its key components.
+
+\subsubsection{Phase 1: Legality}
+The main part of this phase is checking whether it is even legal to optimize a given loop. 
+In addition to the general vectorization legality concepts mentioned in 
+Section~\ref{sec:vec-legal}, LV has a number of extra conditions to check.
+
+Some of the most important legality checks done are:
+\begin{itemize}
+  \item The scalar loop's control flow must not contain indirect branches or multiple backedges.
+  An indirect branch is a branch that jumps to a value evaluated at runtime. A backedge in a loop 
+  is an edge from the loop back to it. 
+  \item Loops with non-linear control flow must be if-convertible. This means any branch or switch 
+  statement must not exit the loop, all instructions of such a block can be executed speculatively 
+  (e.g. via masking)
+  \item Loop-carried dependencies must be vectorizable (see Section~\ref{sec:vec-legal}). In LV 
+  this means explicitly that such a value must \textit{at least} either be an induction variable, a reduction 
+  variable, or a fixed-order recurrence.
+  \item The loop must not contain function calls outside of standard math library calls.
+  \item Types must be vectorizable. Vectorizable types are integers, floating point types, and 
+  pointers.
+  \item The number of runtime checks to guarantee no pointers used by recurrences overlap must not be too high.
+\end{itemize}
+
+One can find a lot more details about these legality checks and their implementations by looking 
+at the LLVM codebase under \texttt{LoopVectorizationLegality::\allowbreak canVectorize}.
+
+\subsubsection{Phase 2: Planning}
+After the Legality Phase has analyzed the scalar loop, initial VPlans are constructed.
+A VPlan is initialized by first creating the H-CFG skeleton building blocks like preheaders, a 
+region block for the vector loop, a middle block with the potential runtime check for the scalar 
+epilogue loop, and a block for the scalar loop. After the H-CFG has been created, the scalar LLVM IR 
+is iterated in topological order and each scalar IR instruction is transformed into a \texttt{VPRecipe}.
+After finishing modeling each scalar IR instruction into the VPlan, a number of VPlan transformations 
+are executed, optimizing the initialized VPlan.
+
+After the initial VPlans have been optimized the best Vectorization Factor is computed. This is done 
+by comparing all generated VPlans for all their possible VFs and choosing the most profitable one. If 
+this step computes a optimal VF of $1$, it means that any form of vectorization is not profitable.
+
+\paragraph{Cost Model}
+Currently LV and VPlan still mostly rely on the old cost model. This legacy cost model derives costs 
+based on the scalar LLVM IR instructions or, if applicable, their vectorized equivalent.
+This is suboptimal for VPlan as VPlans work on \texttt{VPRecipes} abstracted away from the underlying 
+IR. These costs per instructions are then generally just lookup tables influenced by the underlying 
+compile target.
+
+\subsubsection{Phase 3: Execution}
+After the best VPlan has been chosen it is executed (\texttt{LoopVectorizationPlanner::executePlan}).
+For this VPlan some final transformations are made that depend on a specifically chosen VF and UF.
+First a vectorized code skeleton is created in the IR, then the VPlan is executed. Upon execution 
+the H-CFG of the VPlan is walked and each node is executed, which in turn executes all of its 
+\texttt{VPRecipes} in order.
+
+Figure~\ref{fig:inner-loop-vec-cfg} shows the general CFG of a vectorized loop in LLVM and for branches 
+the general branch condition per block.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=0.35\textwidth]{images/inner-loop-vec-loop-cfg-color.png}
+  \caption{The generalized CFG for a vectorized loop with trip count \texttt{N}, 
+  iteration variable \texttt{i}, vectorization factor \texttt{VF}, 
+  and unroll factor \texttt{UF} in LLVM. A VPlan would explicitly model the red-highlighted 
+  region in its H-CFG.}
+  \label{fig:inner-loop-vec-cfg}
+\end{figure}
+
+In case that the epilogue loop could also be further vectorized, it will be in a second 
+pass by creating VPlans for it, finding the best one and executing it.
 
@@
 many cost model-related decisions. However, the transition to a VPlan-based cost model has already 
-started.
+started. This new model would derive costs based on \texttt{VPRecipes} directly instead of the 
+abstracted scalar IR instructions.
 
@@
 
-One such project that also tries to enable auto-vectorization support for inner-, outer-loop, and 
-whole-function vectorization is the region vectorizer from the University Saarland~\cite{rv}.
-This region-vectorizer implements partial control-flow linearization~\cite{10.1145/3192366.3192413}
+One such project that tries to enable auto-vectorization support for inner/outer loops and 
+whole-functions is the region vectorizer (RV) from the University Saarland~\cite{rv}.
+The idea is that all the different vectorization strategies shown in Section~\ref{back:vec-strat}
+can be generalized into a single vectorization approach.
+
+For example, function vectorization can be transformed into loop vectorization by creating a loop 
+around the function's body and vectorizing it and vice versa. 
+The same applies for SLP Vectorization and loop vectorization by unrolling a loop.
+
+RV's goal is to have such a generalized auto-vectorization by vectorizing over \textit{regions}.
+A region is defined as any control flow that has a single entry and one or more non-divergent exits.
+Non-divergent exits mean that starting from the same entry, control flow must not diverge to different
+exits.
+Such a region can then be for example an inner/outer loop, function, or SLP code~\cite{rvintro,rvproposal}.
+
+To achieve this, RV implements partial control-flow linearization~\cite{10.1145/3192366.3192413}
 and a divergence analysis that can handle unstructured CFGs. There has been an effort to partially
@@
 In conclusion, VPlan has been in active development for several years now. When compiling with LLVM, 
-VPlans model initial vectorization approaches, keep track of different optimization decisions, 
-and finally, the best VPlan executes to transform the underlying IR. 
-However, some VPlan-to-VPlan transformations and a VPlan-based cost model must be included before the 
-legacy LV code can be discarded entirely.
+VPlans are used to model different vectorization approaches. They do so by adding a new layer of 
+abstraction over the underlying LLVM IR. This VPlan layer uses \texttt{VPRecipes} to model vectorization 
+specific instructions. This abstraction allows for efficiently exploring many different optimization 
+decisions at once, keeping the underlying implementation flexible and modular. In the end a cost model, 
+currently still working based on LLVM IR instead of \texttt{VPRecipes} is used to compare different VPlans 
+and choose the most optimal one. Finally this chosen VPlan is executed, materializing its \texttt{VPRecipes} 
+back into LLVM IR.
 
-Outer loop vectorization support via VPlan is still experimental, separated from the inner loop
-LV integration, and runs on a divergent experimental path (\texttt{-enable-\allowbreak vplan-\allowbreak native-\allowbreak path}).
-In the near future, development on outer loop vectorization will continue, and merging 
+Outer loop vectorization support via VPlan is still experimental and separated from the inner loop
+LV integration. In the near future, development on outer loop vectorization will continue, and merging 
 it with LV's VPlan path is planned.
@@
 
-\appendix
-
-\section{Appendix: Additional Figures}\label{appendix-figures}
-
-\begin{figure*}
-  \centering
-  \includegraphics[width=\textwidth]{images/inner-loop-vplan-hcfg.png}
-  \caption{Full hierarchical CFG of a VPlan modelling the vectorized version of the loop 
-  depicted in Figure~\ref{fig:inner-loop-vec} as of LLVM 19.1.4.}
-  \label{fig:inner-loop-vplan-hcfg}
-\end{figure*}
-
 \end{document}

### 2024-12-05 paper/paper.tex
paper

@@
   innermost loops. Other areas, such as outer loop vectorization, are also in development.
-  In this paper we give an introduction into VPlan and summarize its latest development results.
+  In this paper, we introduce VPlan and summarize its latest development results.
 \end{abstract}
@@
 This system, however, had quite a few limitations, as the loop vectorizer could only handle
-innermost loops and neither outer loops, complex control flow, nor non-inlined functions. 
+innermost loops and neither outer loops, complex control-flow, nor non-inlined functions. 
 Additionally, while multiple different vectorizations for the same scalar code 
@@
 in a single flexible system capable of optimizing SLPs, inner, and nested loops with complex 
-control flow.
+control-flow.
 
-The auto-vectorization pass would create multiple different VPlans, based on initial legality 
-checks. Each VPlan represents a different vectorization approaches, by modeling the
+The auto-vectorization pass would create multiple different VPlans based on initial legality 
+checks. Each VPlan represents a different vectorization approach, by modeling the
 underlying scalar control flow with its own abstracted vectorized version. Different VPlan-to-VPlan
-transformations can then optimize these abstracted control flows based on a cost model.
-After that, the best VPlan, and with it the best vectorization approach, is chosen by the cost model.
-This VPlan is then executed, by going over its own vectorized control flow representation and
+transformations can then optimize these abstracted control-flows based on a cost model.
+After that, the best VPlan, and with it, the best vectorization approach is chosen by the cost model.
+This VPlan is then executed by going over its own vectorized control flow representation and
 materializing it back into the underlying LLVM IR.
@@
 
-The rest of this paper is organized as follows. We first explain different vectorization
-strategies and vectorization constraints in general and describe the original auto-vectorization
+The rest of this paper is organized as follows. First, we explain different vectorization
+strategies and constraints in general and describe the original auto-vectorization
 approach in LLVM. Section~\ref{sec:vplan} then introduces the new VPlan approach, focusing on
-its most important individual components and gives an outlook on future developments. In 
-Section~\ref{sec:relatedwork} we highlight related work to LLVM's auto-vectorization. Concluding,
-we summarize VPlan, its current state and potential future.
+its most important individual components, and gives an outlook on future developments. In 
+Section~\ref{sec:relatedwork}, we highlight work related to LLVM's auto-vectorization. Concluding,
+we summarize VPlan, its current state, and its potential future.
 
@@
 this call could be replaced with a call to a newly constructed function with the same semantic 
-control flow but operating on multiple values at once through vector registers.
+control-flow but operating on multiple values at once through vector registers.
 
@@
 
-Uncountable loops can typically also not be vectorized, due to their unknown number of iterations.
+Uncountable loops can typically also not be vectorized due to their unknown number of iterations.
 
-Finally some instructions like integer division do inherently not have a vectorized instruction equivalent.
+Finally, some instructions, such as integer division, do inherently not have a vectorized instruction equivalent.
 
@@
 Some vector instructions can be costly on most CPUs. Examples are gather, scatter, shuffle,
-and permutation instructions. Control flow with many non-adjacent memory loads/stores or 
+and permutation instructions. control flow with many non-adjacent memory loads/stores or 
 operations that combine different vector elements, such as reductions, can decrease vectorization 
@@
 and SLPVectorize~\cite{llvmvec,llvmhistorystate}. Auto-Vectorization is part of LLVM's 
-middle-end and runs after module simplification has taken place.
-In general, these vectorizers take in scalar LLVM IR and transform it into optimized vectorized 
+middle-end and runs after module simplification.
+These vectorizers take scalar LLVM IR and transform it into optimized vectorized 
 LLVM IR.
@@
 
-The goal of Intels initial VPlan proposal was to unite any 
-auto-vectorization systems (SLP, Inner-/Outer-Loop, Whole-Function) into a single flexible
-system~\cite{llvmextloopvec,llvmintrvplan,llvmvplanstate}. 
+Intel's initial VPlan proposal aimed to unite any 
+auto-vectorization systems (SLP, Inner-/Outer-Loop, Whole-Function) into a 
+single flexible system~\cite{llvmextloopvec,llvmintrvplan,llvmvplanstate}. 
 This system would use VPlans to represent different 
-vectorization approaches, abstracting away from the underlying scalar loop IR by using its own 
-hierarchical control-flow-graph (H-CFG). This allows for iteratively exploring different 
+vectorization approaches, abstracting away from the underlying scalar loop IR using its own 
+hierarchical control flow graph (H-CFG). This allows iteratively exploring different 
 optimization decisions by performing transformations on this CFG without modifying any LLVM IR.
-Each VPlan would be able to calculate its cost and execute itself, materializing the vectorization 
+Each VPlan could to calculate its cost and execute itself, materializing the vectorization 
 approach back into LLVM IR.
 
-This planned systems is shown in Figure~\ref{fig:vplan-future}. At first, the Legality 
+This planned system is shown in Figure~\ref{fig:vplan-future}. At first, the Legality 
 Phase would check if the scalar code would be legal to vectorize at all, and if so, 
@@
 \subsection{VPlan Structure}
-Before explaining how far this system is already implemented in LLVM we will go over
+Before explaining how far this system has been implemented in LLVM, we will review
 the different elements of a \texttt{VPlan} in more detail.
@@
 The purpose of a single VPlan object is to model a candidate for vectorization. 
-Such a single candidate can potentially represent multiple different vectorized loops in the end. 
-For example, one VPlan might model a general vectorization approach for multiple different 
-Vectorization Factors (VFs) and multiple different Unroll Factors (UFs) if applicable.
+Such a single candidate can potentially represent multiple different vectorized loops. 
+For example, one VPlan might model a general vectorization approach for various 
+Vectorization Factors (VFs) and Unroll Factors (UFs).
 
@@
 Each VPlan stores its own Hierarchical-CFG (H-CFG) modeling the vectorization candidate. This 
-control-flow graph and its components are the main element of a single VPlan.
+control-flow graph and its components are the main elements of a single VPlan.
 The nodes of this H-CFG are one of three types:
@@
 \begin{itemize}
-  \item \texttt{VPBasicBlock} stores a list of \texttt{VPRecipes}, similar to 
+  \item \texttt{VPBasicBlock} stores a list of \texttt{VPRecipes} similar to 
   the LLVM IR's \texttt{BasicBlock} holding a list of IR Instructions.
@@
   \item \texttt{VPIRBasicBlock} is a special \texttt{VPBasicBlock} that wraps an underlying
-  IR \texttt{BasicBlock}. These nodes are useful for modeling the scalar parts around the 
-  final vectorized loops such as the loop preheader and the epilogue loop.
+  IR \texttt{BasicBlock}. These nodes help model the scalar parts around the 
+  final vectorized loops, such as the loop preheader and the epilogue loop.
 \end{itemize}
@@
 The \texttt{VPRecipes} in Figure~\ref{fig:inner-loop-vplan-hcfg-body} have a \texttt{PREFIX}
-infront of them indicating their type. Additionally instructions/recipes of the same color 
+in front of them indicating their type. Additionally instructions/recipes of the same color 
 between Figure~\ref{fig:inner-loop-scalar-ir} and Figure~\ref{fig:inner-loop-vplan-hcfg-body}
 are related. The only recipes in our example VPlan CFG that do not semantically model any 
-instructions found in the scalar IR are from the \texttt{middle.block} (purple). This block 
-is used to decide if the scalar epilogue loop is to be executed following the vectorized loop.
+instructions found in the scalar IR are from the \texttt{middle.block} (purple). 
+This block decides if the scalar epilogue loop will be executed following the vectorized loop.
 
@@
 In its current state, VPlan is already deeply integrated into LoopVectorize (LV) and is
-used to model, optimize and execute vectorization of innermost loops~\cite{llvmvplan,llvmintrvplan,llvmvplanupdate}. 
-While outer loop vectorization via VPlan has already seen quite some development, any 
-outer loop vectorization features are still experimental and need to be explicitly enabled 
+used to model, optimize, and execute vectorization of innermost loops~\cite{llvmvplan,llvmintrvplan,llvmvplanupdate}. 
+While outer loop vectorization via VPlan has already seen some development, any 
+outer loop vectorization features are still experimental and must be explicitly enabled 
 by passing the \texttt{-enable-\allowbreak vplan-\allowbreak native-\allowbreak path} 
@@
 In addition to the general vectorization legality concepts mentioned in 
-Section~\ref{sec:vec-legal}, LV has a number of extra conditions to check.
+Section~\ref{sec:vec-legal}, LV has a several extra conditions to check.
 
@@
 \begin{itemize}
-  \item The scalar loop's control flow must not contain indirect branches or multiple backedges.
-  An indirect branch is a branch that jumps to a value evaluated at runtime. A backedge in a loop 
+  \item The scalar loop's control flow must not contain indirect branches or multiple back-edges.
+  An indirect branch is a branch that jumps to a value evaluated at runtime. A back-edge in a loop 
   is an edge from the loop back to it. 
   \item Loops with non-linear control flow must be if-convertible. This means any branch or switch 
-  statement must not exit the loop, all instructions of such a block can be executed speculatively 
+  statement must not exit the loop. All instructions of such a block can be executed speculatively 
   (e.g. via masking)
@@
   \item The loop must not contain function calls outside of standard math library calls.
-  \item Types must be vectorizable. Vectorizable types are integers, floating point types, and 
-  pointers.
   \item The number of runtime checks to guarantee no pointers used by recurrences overlap must not be too high.
@@
 
-One can find a lot more details about these legality checks and their implementations by looking 
+One can find more details about these legality checks and their implementations by looking 
 at the LLVM codebase under \texttt{LoopVectorizationLegality::\allowbreak canVectorize}.
@@
 After the Legality Phase has analyzed the scalar loop, initial VPlans are constructed.
-A VPlan is initialized by first creating the H-CFG skeleton building blocks like preheaders, a 
+A VPlan is initialized by creating the H-CFG skeleton building blocks like preheaders, a 
 region block for the vector loop, a middle block with the potential runtime check for the scalar 
 epilogue loop, and a block for the scalar loop. After the H-CFG has been created, the scalar LLVM IR 
-is iterated in topological order and each scalar IR instruction is transformed into a \texttt{VPRecipe}.
-After finishing modeling each scalar IR instruction into the VPlan, a number of VPlan transformations 
+is iterated in topological order, and each scalar IR instruction is transformed into a \texttt{VPRecipe}.
+After modeling each scalar IR instruction with a \texttt{VPRecipe}, several VPlan transformations 
 are executed, optimizing the initialized VPlan.
 
-After the initial VPlans have been optimized the best Vectorization Factor is computed. This is done 
+After the initial VPlans have been optimized, the best Vectorization Factor is computed. This is done 
 by comparing all generated VPlans for all their possible VFs and choosing the most profitable one. If 
-this step computes a optimal VF of $1$, it means that any form of vectorization is not profitable.
+this step computes an optimal VF of $1$, any vectorization is not profitable compared to the 
+scalar version.
 
 \paragraph{Cost Model}
-Currently LV and VPlan still mostly rely on the old cost model. This legacy cost model derives costs 
+Currently, LV and VPlan still mostly rely on the old cost model. This legacy cost model derives costs 
 based on the scalar LLVM IR instructions or, if applicable, their vectorized equivalent.
 This is suboptimal for VPlan as VPlans work on \texttt{VPRecipes} abstracted away from the underlying 
-IR. These costs per instructions are then generally just lookup tables influenced by the underlying 
+IR. These costs per instruction are then generally just lookup tables influenced by the underlying 
 compile target.
@@
 \subsubsection{Phase 3: Execution}
-After the best VPlan has been chosen it is executed (\texttt{LoopVectorizationPlanner::executePlan}).
-For this VPlan some final transformations are made that depend on a specifically chosen VF and UF.
-First a vectorized code skeleton is created in the IR, then the VPlan is executed. Upon execution 
-the H-CFG of the VPlan is walked and each node is executed, which in turn executes all of its 
-\texttt{VPRecipes} in order.
+After the best VPlan has been chosen, it is executed (\texttt{LoopVectorizationPlanner::executePlan}).
+For this VPlan, some final transformations are run depending on a specifically chosen VF and UF.
+First, a vectorized code skeleton is created in the IR, and then the VPlan is executed. 
+Upon execution, the H-CFG of the VPlan is walked, and each node is executed, which executes all 
+of its \texttt{VPRecipes} in order.
 
-Figure~\ref{fig:inner-loop-vec-cfg} shows the general CFG of a vectorized loop in LLVM and for branches 
+Figure~\ref{fig:inner-loop-vec-cfg} shows the general CFG of a vectorized loop in LLVM and, for branches, 
 the general branch condition per block.
@@
 
-In case that the epilogue loop could also be further vectorized, it will be in a second 
-pass by creating VPlans for it, finding the best one and executing it.
+If the epilogue loop could be further vectorized, it would be in a second pass by creating VPlans, 
+finding the best one, and executing it.
 
@@
 around the function's body and vectorizing it and vice versa. 
-The same applies for SLP Vectorization and loop vectorization by unrolling a loop.
+The same applies to SLP Vectorization and loop vectorization by unrolling a loop.
 
-RV's goal is to have such a generalized auto-vectorization by vectorizing over \textit{regions}.
-A region is defined as any control flow that has a single entry and one or more non-divergent exits.
-Non-divergent exits mean that starting from the same entry, control flow must not diverge to different
-exits.
-Such a region can then be for example an inner/outer loop, function, or SLP code~\cite{rvintro,rvproposal}.
+RV aims to have such a generalized auto-vectorization by vectorizing over \textit{regions}.
+A region is defined as any control flow with a single entry and one or more non-divergent exits.
+Non-divergent exits mean that control flow must not diverge to different exits starting from the 
+same entry.
+Such a region can then be, for example, an inner/outer loop, function, or SLP code~\cite{rvintro,rvproposal}.
 
-To achieve this, RV implements partial control-flow linearization~\cite{10.1145/3192366.3192413}
+To achieve this, RV implements partial control flow linearization~\cite{10.1145/3192366.3192413}
 and a divergence analysis that can handle unstructured CFGs. There has been an effort to partially
-implement this control-flow linearization and divergence analysis into LLVM's 
+implement this control flow linearization and divergence analysis into LLVM's 
 upstream~\cite{rvproposal,rvproposaldep}.
@@
 VPlans are used to model different vectorization approaches. They do so by adding a new layer of 
-abstraction over the underlying LLVM IR. This VPlan layer uses \texttt{VPRecipes} to model vectorization 
-specific instructions. This abstraction allows for efficiently exploring many different optimization 
-decisions at once, keeping the underlying implementation flexible and modular. In the end a cost model, 
-currently still working based on LLVM IR instead of \texttt{VPRecipes} is used to compare different VPlans 
-and choose the most optimal one. Finally this chosen VPlan is executed, materializing its \texttt{VPRecipes} 
+abstraction over the underlying LLVM IR. This VPlan layer uses \texttt{VPRecipes} to model 
+vectorization-specific instructions. 
+This abstraction allows for efficiently exploring many optimization decisions simultaneously, 
+keeping the underlying implementation flexible and modular. In the end, a cost model currently 
+working based on LLVM IR instead of \texttt{VPRecipes} is used to compare different VPlans 
+and choose the most optimal one. Finally, this chosen VPlan is executed, materializing its \texttt{VPRecipes} 
 back into LLVM IR.

### 2025-01-26 paper/paper.tex
paper

@@
 
-Uncountable loops can typically also not be vectorized due to their unknown number of iterations.
+Uncountable loops, meaning loops where the total iteration count cannot be determined before the 
+loop is executed, can often not be vectorized. This is because their exit condition usually depends on 
+the loop's body itself, making it impossible to predetermine how many vector lanes are needed for the next 
+loop iteration when executing.
 
-Finally, some instructions, such as integer division, do inherently not have a vectorized instruction equivalent.
+Finally, some instructions do generally not have a direct vectorized 
+instruction equivalent. For example Integer division, commonly implemented using the 
+\textit{Newton-Raphson method}, is already a complex and iterative operation 
+on modern CPUs. Applying this algorithm 
+to multiple data elements in a single vector register would result in even more implementation complexity 
+and is thus considered not worthwhile in most cases.
+Workarounds for some specialized cases do exist~\cite{vecintdiv}.
 
@@
 
-Some vector instructions can be costly on most CPUs. Examples are gather, scatter, shuffle,
-and permutation instructions. control flow with many non-adjacent memory loads/stores or 
-operations that combine different vector elements, such as reductions, can decrease vectorization 
-performance.
+The actual performance of a vector instruction also heavily depends on the ISA and its 
+hardware implementation. 
+For example, gather, scatter, 
+shuffle, and permutation instructions are typically slow. Control flow with many 
+non-adjacent memory accesses or 
+operations that combine different vector elements from the same register (horizontal reductions) can thus 
+decrease vectorization performance.
 
@@
 checks might be inserted throughout the vectorized control flow to handle, for example, pointers 
-to overlapping memory regions, in which case the scalar version of the loop is executed.
+to overlapping memory regions, in which case the scalar version of the loop must be executed due to 
+unclear data dependencies.
 
@@
 Each VPlan stores its own Hierarchical-CFG (H-CFG) modeling the vectorization candidate. This 
-control-flow graph and its components are the main elements of a single VPlan.
-The nodes of this H-CFG are one of three types:
+control-flow graph and its components are the main elements of a single VPlan. It is hierarchical
+in that a single node of this graph can be a complete subgraph, representing more complex nested
+behavior.
+Specifically, any node of this H-CFG is one of three types:
 
@@
 in Figure~\ref{fig:inner-loop-vec}.
+% TODO: Explain contents of both figures better
 
@@
   \item \texttt{VPReplicateRecipe} will replicate one IR instruction either once (\texttt{CLONE})
-  or multiple times (\texttt{REPLICATE}) into the vectorized IR when executed.
+  or VF * UF times (\texttt{REPLICATE}) into the vectorized IR when executed.
   \item \texttt{VPWidenRecipe} will \texttt{WIDEN} one IR instruction into a
@@
   \texttt{EMIT} one or more new IR instructions.
+  % TODO: Explain some instructions more in detail (e.g. CANONICAL-INDUCTION, SCALAR-STEPS)
 \end{itemize}
@@
 
-\subsection{State of VPlan (LLVM 19.1.4)}
+% TODO: VPlan Transformations example
+
+\section{State of VPlan (LLVM 19.1.4)}
 \emergencystretch 6em
@@
 
-\subsubsection{Phase 1: Legality}
+\subsection{Phase 1: Legality}
 The main part of this phase is checking whether it is even legal to optimize a given loop. 
@@
 
-\subsubsection{Phase 2: Planning}
+\subsection{Phase 2: Planning}
+% TODO: Connect better to previous phase (expand a bit on how the legality artifacts are used when planning)
 After the Legality Phase has analyzed the scalar loop, initial VPlans are constructed.
@@
 compile target.
+% TODO: Expand on this and explain a bit more
 
-\subsubsection{Phase 3: Execution}
+\subsection{Phase 3: Execution}
 After the best VPlan has been chosen, it is executed (\texttt{LoopVectorizationPlanner::executePlan}).
@@
 First, a vectorized code skeleton is created in the IR, and then the VPlan is executed. 
-Upon execution, the H-CFG of the VPlan is walked, and each node is executed, which executes all 
+Upon execution, the H-CFG of the VPlan is walked, and each node is executed, which in turn executes all 
 of its \texttt{VPRecipes} in order.
+
+% TODO: How exactly is the translation from VP recipe back to LLVM IR working? Where does the target
+%       backend come in to decide what actual instruction?
 
@@
   iteration variable \texttt{i}, vectorization factor \texttt{VF}, 
-  and unroll factor \texttt{UF} in LLVM. A VPlan would explicitly model the red-highlighted 
-  region in its H-CFG.}
+  and unroll factor \texttt{UF} in LLVM. A VPlan would so far explicitly model the red-highlighted 
+  region in its H-CFG. If a node branches, the condition is stated in the body, and the two possible 
+  path edges are marked with T (True) and F (False).}
   \label{fig:inner-loop-vec-cfg}
@@
 
-\subsection{Future of VPlan}
+\section{Future of VPlan}
 Many of the VPlan internals and surrounding systems are still in active development. To stay up-to-date 
@@
 
-\subsubsection{VPlan and Integration with LV}
+\subsection{VPlan and Integration with LV}
 Setting up and integrating the VPlan structure into LV has seen tremendous effort over the last 
@@
 
-\subsubsection{Outer Loop Vectorization}
+\subsection{Outer Loop Vectorization}
 Development around outer loop vectorization has slowed as the focus shifted to VPlan internals 
@@
 
-\subsubsection{Merging with SLP Vectorize}
+\subsection{Merging with SLP Vectorize}
 Merging the existing SLP Vectorizer with the current VPlan infrastructure is in the early planning phase.
 
-\subsubsection{Function Vectorization}
+\subsection{Function Vectorization}
 There is currently no active development for whole-function vectorization in the
@@
 
+% TODO: Dependency Analysis (SCEV Expressions)
+%         - https://www.npopov.com/2023/10/03/LLVM-Scalar-evolution.html
+%         - https://www.youtube.com/watch?v=AmjliNp0_00
+% TODO: Polly and its use for loop transformations affecting data dependencies
+% TODO: Comparisons LLVM with GCC
+%         - Tail loop folding (https://github.com/llvm/llvm-project/issues/123069)
+%         - Another benchmark: https://www.spec.org/cpu2017/Docs/benchmarks/525.x264_r.html
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
+% TODO: Write about potential future work benchmarking
+%       Loop Code Benchmark collection
+%       Quick note about collaborative development (example stability & miscompile policy)
+%       LLVM Loop Optimization as one of the most active subareas of LLVM development
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2025-01-27 paper/paper.tex
paper

@@
 \label{back:vec-strat}
-In general, vectorization techniques can be separated into different categories:
+In general, vectorization strategies can be separated into different categories:
 
@@
 \subsection{VPlan Structure}
+\emergencystretch 2em
 Before explaining how far this system has been implemented in LLVM, we will review
@@
 
-Figure~\ref{fig:inner-loop-scalar-ir} shows the loop part of the scalar IR for the code example
-of Figure~\ref{fig:inner-loop-vec}. Comparatively, Figure~\ref{fig:inner-loop-vplan-hcfg-body}
-shows the loop part of the H-CFG of a VPlan that models the vectorization approach described
-in Figure~\ref{fig:inner-loop-vec}.
-% TODO: Explain contents of both figures better
-
 \begin{figure}
@@
   \includegraphics[width=0.475\textwidth]{images/inner-loop-vplan-hcfg-loop-body-color.png}
-  \caption{The vectorized loop section of the hierarchical CFG of a VPlan modeling the 
+  \caption{The vectorized loop part of the hierarchical CFG of a VPlan modeling the 
   conditional loop depicted in Figure~\ref{fig:inner-loop-vec}. 
@@
   Some predefined values are
-  \texttt{vp<\%0>} = VF * UF; \texttt{vp<\%1>} = vector-trip-count; \texttt{vp<\%2>} = original trip-count.}
+  \texttt{vp<\%0>} = VF * UF; \texttt{vp<\%1>} = vector trip count; \texttt{vp<\%2>} = original trip count.}
   \label{fig:inner-loop-vplan-hcfg-body}
 \end{figure}
+
+Figure~\ref{fig:inner-loop-scalar-ir} shows the loop part of the scalar IR for the code example
+of Figure~\ref{fig:inner-loop-vec}. Comparatively, Figure~\ref{fig:inner-loop-vplan-hcfg-body}
+shows the vectorized loop part of the H-CFG of a VPlan that models the vectorization approach described
+in Figure~\ref{fig:inner-loop-vec}. Note that the full H-CFG of this VPlan consists of many more nodes 
+to handle legality checks (coming before the vectorized loop) or the entire scalar epilogue loop 
+(coming after the vectorized loop).
+
+The H-CFG in Figure~\ref{fig:inner-loop-vplan-hcfg-body} consists of 3 nodes on the highest layer. 
+\texttt{vector.preheader} and \texttt{middle.block} 
+are simple \texttt{VPBasicBlocks} surrounding the main vector loop node. This \texttt{vector loop} node 
+is a \texttt{VPRegionBlock} that does 
+not replicate its contents (indicated by \texttt{<x1>}; otherwise, it would be \texttt{<xVFxUF>}).
+
+Inside \texttt{vector loop} is only a single \texttt{VPBasicBlock} node \texttt{vector.body}. This block 
+models the entire vectorized loop via \texttt{VPRecipes}.
 
@@
 \texttt{VPRecipes} can be seen as instructions of an IR specialized for auto-vectorization. 
-A single \texttt{VPRecipe} inside a VPlan is based on zero or more LLVM IR instructions of the 
-underlying scalar IR and will materialize into one or more LLVM IR instructions when executed.
+A single \texttt{VPRecipe} inside a VPlan is based on zero or more LLVM IR instructions 
+of the underlying scalar IR, called ingredients, and will materialize into one or more LLVM IR 
+instructions when executed.
 
-There are many different types of \texttt{VPRecipes}, most notably:
+There are currently 40 different final types of \texttt{VPRecipes}. The types used 
+within Figure~\ref{fig:inner-loop-vplan-hcfg-body} and the most important ones are:
 \begin{itemize}
@@
   \texttt{EMIT} one or more new IR instructions.
-  % TODO: Explain some instructions more in detail (e.g. CANONICAL-INDUCTION, SCALAR-STEPS)
+  \item \texttt{VPHeaderPHIRecipe} is a sub-type that models any induction, such as first-order 
+  recurrence, pointer induction, or reduction as a phi node.
+  \item \texttt{VPCanonicalIVPHIRecipe} derives from \texttt{VPHeaderPHIRecipe} and represents a 
+  \texttt{CANONICAL-INDUCTION} variable via a \textit{scalar} value.
+  \item \texttt{VPScalarIVStepsRecipe} models the individual offsets (\texttt{SCALAR-STEPS}) per 
+  vector-lane from a common Integer or Floating-Point induction value.
 \end{itemize}
 
-The \texttt{VPRecipes} in Figure~\ref{fig:inner-loop-vplan-hcfg-body} have a \texttt{PREFIX}
-in front of them indicating their type. Additionally instructions/recipes of the same color 
+Most of the \texttt{VPRecipes} in Figure~\ref{fig:inner-loop-vplan-hcfg-body} have capitalized 
+\texttt{HINTS} indicating their type. Additionally instructions/recipes of the same color 
 between Figure~\ref{fig:inner-loop-scalar-ir} and Figure~\ref{fig:inner-loop-vplan-hcfg-body}
-are related. The only recipes in our example VPlan CFG that do not semantically model any 
+are related in their semantical effect.
+
+The only recipes in our example VPlan CFG that do not semantically model any 
 instructions found in the scalar IR are from the \texttt{middle.block} (purple). 
-This block decides if the scalar epilogue loop will be executed following the vectorized loop.
+This block decides if the scalar epilogue loop is executed after the vectorized loop by 
+comparing the vector trip count to the original loop trip count. This additional vector trip count is 
+needed as the original loop trip count might not be a perfect multiple of VF * UF, the number of elements 
+processed in a single iteration.
+
+The red recipes generally handle the induction variable \texttt{i} and the termination of the 
+vectorized loop. \texttt{vp<\%3>} is this loop's scalar induction variable, which goes from 0 to 
+\texttt{vp<\%1>}, the vector trip count. The per vector-lane offset's step size is 1 (\texttt{ir<1>}), 
+meaning that each 
+vector-lane offset from the base induction variable is one greater than the previous lane (we access the 
+memory consecutively).
+
+The blue recipes model the condition guarding the addition. \texttt{vp<\%6>} holds the mask, which 
+results from this condition and is used for the guarded memory operations.
+
+Finally, the green recipes represent the addition. First, the data from both \texttt{y} and \texttt{x} 
+gets loaded (guarded by the compare mask), then added together, and then stored back into 
+\texttt{x} (again, guarded by the compare mask).
 
@@
 
-% TODO: VPlan Transformations example
+Such transformations include for example:
+\begin{itemize}
+  \item Simplifying \texttt{VPRecipes} by removing blend operations with only one unique input, removing 
+  redundant type casts, and simplifying logical patterns.
+  \item Removing dead \texttt{VPRecipes} whose values have no users.
+  \item Moving loop-invariant \texttt{VPRecipes} out of the vector loop region. A \texttt{VPRecipe} is 
+  considered loop-invariant if it does not have any side effects, does not read from memory, is not a 
+  phi node, and all its operands are defined outside of the vector loop region.
+\end{itemize}
 
 \section{State of VPlan (LLVM 19.1.4)}
-\emergencystretch 6em
+\emergencystretch 3em
 In its current state, VPlan is already deeply integrated into LoopVectorize (LV) and is
@@
 To better understand how VPlan is already being used in LLVM, we will now follow along the 
-LoopVectorize pass and summarize its key components.
+\texttt{LoopVectorizePass} and summarize its key components.
 
 \subsection{Phase 1: Legality}
-The main part of this phase is checking whether it is even legal to optimize a given loop. 
+The main part of this first phase is checking whether it is even legal to vectorize a given loop. 
 In addition to the general vectorization legality concepts mentioned in 
@@
 
+Additionally, during this phase, all recurrences of the loop, such as reductions or inductions, are analyzed,
+and their phi nodes are stored in a set.
+
 One can find more details about these legality checks and their implementations by looking 
@@
 \subsection{Phase 2: Planning}
-% TODO: Connect better to previous phase (expand a bit on how the legality artifacts are used when planning)
-After the Legality Phase has analyzed the scalar loop, initial VPlans are constructed.
+After the Legality Phase has analyzed the scalar loop and it is deemed legal to vectorize the 
+\texttt{LoopVectorizationCostModel} and the \texttt{LoopVectorizationPlanner} are initialized.
+The latter one then constructs the initial VPlans based on the properties gathered during the legal phase
+and the cost model initialization, such as the recurrence variables or the maximum viable vectorization 
+factor.
+
 A VPlan is initialized by creating the H-CFG skeleton building blocks like preheaders, a 
@@
 \paragraph{Cost Model}
+\emergencystretch 4em
 Currently, LV and VPlan still mostly rely on the old cost model. This legacy cost model derives costs 
@@
 This is suboptimal for VPlan as VPlans work on \texttt{VPRecipes} abstracted away from the underlying 
-IR. These costs per instruction are then generally just lookup tables influenced by the underlying 
-compile target.
-% TODO: Expand on this and explain a bit more
+IR. These costs per instruction are then generally lookup tables influenced by the chosen target 
+architecture.
+
+The before-mentioned \texttt{LoopVectorizationCostModel} is a remainder of the old \texttt{LoopVectorize} 
+implementation before VPlan. It is a single object that computes different costs based on the scalar loop 
+IR without any relation to different VPlans. For example, it stores the instructions that will remain scalar 
+after vectorization or the LLVM values for which the cost can 
+be ignored. The latter includes, for instance, loop invariant operations that would be hoisted out of the loop 
+or trivially dead instructions.
+
+Additionally, each \texttt{VPRecipe} has a function to calculate its own cost. For most \texttt{VPRecipes} 
+that reference one or more LLVM IR ingredients, these are then used to calculate the cost of the recipe.
+On the other hand, \texttt{VPRecipes} that do not directly model an underlying LLVM scalar IR instruction, 
+such as \texttt{VPInstruction}, can generally not compute their own cost and thus return a cost of zero 
+for now.
+
+For simple innermost loops with few vectorization approaches,
+relying on the old cost model might not have too much of a performance impact. 
+However, once more VPlan transformations and complex
+outer/hybrid loop vectorization are implemented, this new cost model would be an important
+detail of the VPlan system to reliably choose the most optimal vectorization approach.
 
@@
 For this VPlan, some final transformations are run depending on a specifically chosen VF and UF.
-First, a vectorized code skeleton is created in the IR, and then the VPlan is executed. 
+Then a vectorized code skeleton is created in the IR, and the VPlan executed. 
 Upon execution, the H-CFG of the VPlan is walked, and each node is executed, which in turn executes all 
-of its \texttt{VPRecipes} in order.
+of its \texttt{VPRecipes} in order. Each \texttt{VPRecipe} produces its respective output IR.
 
-% TODO: How exactly is the translation from VP recipe back to LLVM IR working? Where does the target
-%       backend come in to decide what actual instruction?
-
-Figure~\ref{fig:inner-loop-vec-cfg} shows the general CFG of a vectorized loop in LLVM and, for branches, 
-the general branch condition per block.
+A \texttt{VPTransformState} object is passed along to store information 
+needed for generating the output IR, such as the chosen vectorization factor or generated values that were 
+used by later \texttt{VPRecipes}.
 
@@
 
-If the epilogue loop could be further vectorized, it would be in a second pass by creating VPlans, 
-finding the best one, and executing it.
+While Figure~\ref{fig:inner-loop-vec-cfg} shows the general CFG of a vectorized loop in LLVM, some 
+nodes can also be omitted in special cases.
+
+For example, if all the pointers used inside the function are marked as \texttt{noalias}, the compiler 
+is guaranteed that they will not overlap and the \texttt{vector.memcheck} block will be omitted.
+While the C language and most modern C++ compilers have an explicit \texttt{restrict} keyword for this,
+Rust automatically applies the \texttt{noalias} attribute to safe references when compiling via LLVM.
+
+Additionally, before the initial VPlans are constructed at the beginning of Phase 2, it is also checked if 
+the scalar epilogue loop could be folded into the vectorized loop version itself by increasing the vector 
+trip count over the actual loop trip count and masking all operations. This tail-folding approach in LLVM 
+has not yet been enabled by default.
+
+Otherwise, if the scalar epilogue loop could be vectorized, it would be in a second loop-vectorize 
+pass by, again, creating VPlans, finding the best one, and executing it.
 
@@
 started. This new model would derive costs based on \texttt{VPRecipes} directly instead of the 
-abstracted scalar IR instructions.
+abstracted scalar IR instructions. Besides that, the community also focuses on enabling tail folding by
+default, which would lead to a significant performance increase in some instances.
 
@@
 %       LLVM Loop Optimization as one of the most active subareas of LLVM development
+% TODO: "Also, it would be great to have the main problem and paper’s aim to be re-stated in the summary section for the first reading pass."
+%       "summary lacks the problem statement and the big picture"
 

### 2025-01-27 paper/paper.tex
paper

@@
 \label{sec:summary}
-In conclusion, VPlan has been in active development for several years now. When compiling with LLVM, 
-VPlans are used to model different vectorization approaches. They do so by adding a new layer of 
-abstraction over the underlying LLVM IR. This VPlan layer uses \texttt{VPRecipes} to model 
-vectorization-specific instructions. 
-This abstraction allows for efficiently exploring many optimization decisions simultaneously, 
-keeping the underlying implementation flexible and modular. In the end, a cost model currently 
+Compilers have been using auto-vectorization to help developers optimize their code 
+across platforms by making the best possible use of a target's SIMD capabilities. Over the years, 
+LLVM's original auto-vectorization system evolved into an inflexible pipeline, making it hard 
+to adapt to new features like outer-loop vectorization.
+
+As a solution, Intel proposed a new modular system called VPlan.
+By now, VPlan has been in active development for several years. 
+This system would use different VPlan objects to efficiently 
+model and explore different vectorization approaches simultaneously.
+It adds a new layer using \texttt{VPRecipes} to model 
+vectorization-specific instructions over the underlying LLVM IR. 
+This abstraction allows for efficiently exploring different optimization decisions simultaneously, 
+keeping the underlying implementation flexible and modular. In the end, a cost model currently still 
 working based on LLVM IR instead of \texttt{VPRecipes} is used to compare different VPlans 
@@
 %       LLVM Loop Optimization as one of the most active subareas of LLVM development
-% TODO: "Also, it would be great to have the main problem and paper’s aim to be re-stated in the summary section for the first reading pass."
-%       "summary lacks the problem statement and the big picture"
 

### 2025-01-28 paper/paper.tex
paper

@@
 \usepackage{todonotes}
+\usepackage{flushend}
+\usepackage{xurl}
 \usepackage[babel]{csquotes}
@@
 (2) As the loop trip count might not be a perfect multiple of the Vectorization Factor, a scalar
-epilogue/remainder of the loop might be used to handle the tail iterations. (3) Various runtime
+epilogue/tail of the loop might be used to handle the remaining loop iterations. (3) Various runtime
 checks might be inserted throughout the vectorized control flow to handle, for example, pointers 
@@
 Intel's initial VPlan proposal aimed to unite any 
-auto-vectorization systems (SLP, Inner-/Outer-Loop, Whole-Function) into a 
+auto-vectorization systems (SLP, Inner/Outer-Loop, Whole-Function) into a 
 single flexible system~\cite{llvmextloopvec,llvmintrvplan,llvmvplanstate}. 
@@
 \subsection{VPlan Structure}
-\emergencystretch 2em
 Before explaining how far this system has been implemented in LLVM, we will review
@@
 \paragraph{H-CFG}
+\emergencystretch 1em
+
 This is achieved by introducing a new layer of abstraction from the underlying loop's scalar IR.
@@
 
-Inside \texttt{vector loop} is only a single \texttt{VPBasicBlock} node \texttt{vector.body}. This block 
+Inside the \texttt{vector loop} region is only a single \texttt{VPBasicBlock} node \texttt{vector.body}. This block 
 models the entire vectorized loop via \texttt{VPRecipes}.
@@
 \paragraph{Recipes}
+\emergencystretch 2em
 
@@
 \section{State of VPlan (LLVM 19.1.4)}
-\emergencystretch 3em
 In its current state, VPlan is already deeply integrated into LoopVectorize (LV) and is
@@
 
-One can find more details about these legality checks and their implementations by looking 
-at the LLVM codebase under \texttt{LoopVectorizationLegality::\allowbreak canVectorize}.
-
 \subsection{Phase 2: Planning}
@@
 \paragraph{Cost Model}
-\emergencystretch 4em
 Currently, LV and VPlan still mostly rely on the old cost model. This legacy cost model derives costs 
@@
 
-The before-mentioned \texttt{LoopVectorizationCostModel} is a remainder of the old \texttt{LoopVectorize} 
-implementation before VPlan. It is a single object that computes different costs based on the scalar loop 
+A remainder of the old \texttt{LoopVectorize} implementation before VPlan is the 
+before-mentioned \texttt{LoopVectorizationCostModel}. It is a single object that computes different costs based on the scalar loop 
 IR without any relation to different VPlans. For example, it stores the instructions that will remain scalar 
@@
 A \texttt{VPTransformState} object is passed along to store information 
-needed for generating the output IR, such as the chosen vectorization factor or generated values that were 
+needed for generating the output IR, such as the chosen Vectorization Factor or generated values that were 
 used by later \texttt{VPRecipes}.
@@
   \caption{The generalized CFG for a vectorized loop with trip count \texttt{N}, 
-  iteration variable \texttt{i}, vectorization factor \texttt{VF}, 
-  and unroll factor \texttt{UF} in LLVM. A VPlan would so far explicitly model the red-highlighted 
+  iteration variable \texttt{i}, Vectorization Factor \texttt{VF}, 
+  and Unroll Factor \texttt{UF} in LLVM. A VPlan would so far explicitly model the red-highlighted 
   region in its H-CFG. If a node branches, the condition is stated in the body, and the two possible 
@@
 the scalar epilogue loop could be folded into the vectorized loop version itself by increasing the vector 
-trip count over the actual loop trip count and masking all operations. This tail-folding approach in LLVM 
+trip count over the actual loop trip count and masking all operations. This tail-loop folding approach in LLVM 
 has not yet been enabled by default.
@@
 started. This new model would derive costs based on \texttt{VPRecipes} directly instead of the 
-abstracted scalar IR instructions. Besides that, the community also focuses on enabling tail folding by
+abstracted scalar IR instructions. Besides that, the community also focuses on enabling tail-loop folding by
 default, which would lead to a significant performance increase in some instances.
@@
 \subsection{Function Vectorization}
-There is currently no active development for whole-function vectorization in the
+There seems to be currently no active development on whole-function vectorization in the
 llvm-project's upstream.
@@
 \label{sec:relatedwork}
+\subsection{Region Vectorizers}
 Next to the main llvm-projects upstream, many projects are using or extending the LLVM toolchain.
@@
 
-To achieve this, RV implements partial control flow linearization~\cite{10.1145/3192366.3192413}
+To achieve this, RV implements partial control flow linearization~\cite{Moll_2021,10.1145/3192366.3192413}
 and a divergence analysis that can handle unstructured CFGs. There has been an effort to partially
@@
 
-% TODO: Dependency Analysis (SCEV Expressions)
-%         - https://www.npopov.com/2023/10/03/LLVM-Scalar-evolution.html
-%         - https://www.youtube.com/watch?v=AmjliNp0_00
-% TODO: Polly and its use for loop transformations affecting data dependencies
-% TODO: Comparisons LLVM with GCC
-%         - Tail loop folding (https://github.com/llvm/llvm-project/issues/123069)
-%         - Another benchmark: https://www.spec.org/cpu2017/Docs/benchmarks/525.x264_r.html
+In a later paper, Moll et al. further generalized this region vectorizer to be able to vectorize 
+nested loops in a hybrid/multi-dimensional way, pushing optimization capabilities even 
+further~\cite{10.1145/3303117.3306172}.
+
+\subsection{Comparison: GCC and LLVM}
+As another popular compiler, GCC has implemented and consistently improved its auto-vectorization 
+over the years~\cite{autovecoptgcc,gccvecsum}. Compared to LLVM, GCC also supports 
+vectorization of outer loops and tail-loop folding by default.
+
+In recent years, different research has been done about how compilers, especially GCC and 
+LLVM, compare in terms of auto-vectorization.
+
+Feng et al. compared multiple versions of GCC, LLVM, and Intel's ICC compiler in auto-vectorizing the 
+TSVC benchmark set targeting x86-64~\cite{feng2021evaluation}. They found that in terms of compilation time, GCC 9
+has much less overhead ($\approx4\%$) when auto-vectorization is enabled compared to 
+LLVM 9 ($\approx24\%$). In both cases, LLVM 9 takes twice as long to compile than GCC 9. 
+ICC 19 sits right between GCC 9 and LLVM 9 regarding compilation time.
+
+While in terms of runtime, GCC 9 and LLVM 9 take around 
+the same time with auto-vectorization disabled, enabling it improves GCC 9's performance by around $44\%$, 
+while for LLVM 9 it is only $12\%$. ICC 19 has a better runtime than GCC 9 and LLVM 9 in both the 
+non-vectorized and the auto-vectorized case.
+
+The authors also investigated the number of successfully auto-vectorized code routines inside the TSVC 
+benchmark divided into different vectorization categories. Out of a total of 151 different routines, GCC 9 
+vectorized 68, LLVM 9 vectorized 49, and ICC 19 did 98.
+
+Feng et al. concluded that auto-vectorizers were mostly limited by analysis capabilities 
+needed to reason about specific optimization approaches and inaccurate cost models.
+
+While LLVM 9 performs poorly in this paper, it is important to notice that at the time of LLVM 9, VPlan was relatively 
+early into development ($\approx3$ years) and was still missing 
+support for a lot of commonly used vectorization idioms such as reductions, for example.
+
+In a more recent auto-vectorization evaluation comparing GCC 14 with LLVM 17 on the similar TSVC 2 
+benchmark targeting RISC-V, the author investigated the produced vectorized code 
+and drew comparisons to similar papers~\cite{gccllvmveccomp}. The SVE and RISC-V 
+vector extensions are  
+vector-length agnostic, 
+meaning that, unlike SSE or AVX, vector instructions are not tied to specific vector register sizes. 
+This makes the code more compatible and scalable across different underlying hardware.
+
+In the paper, the author concludes that the achieved performance uplifts by auto-vectorization are very 
+similar between GCC 14 and LLVM 17, with GCC only being marginally better. LLVM 17 tends to favor 
+generating vector code targeting specific vector sizes. However, it is unclear whether this is an 
+actual drawback.
+
+Finally, as mentioned before, the LLVM auto-vectorization community is currently working on 
+enabling tail-loop folding by default after noticing a $\approx24\%$ performance difference between the latest 
+versions of GCC and LLVM in a SPEC CPU 2017 benchmark targeting RISC-V~\cite{riscvtailfolding}.
 
@@
 LLVM's original auto-vectorization system evolved into an inflexible pipeline, making it hard 
-to adapt to new features like outer-loop vectorization.
+to adapt to new features like outer loop vectorization.
 
@@
 LV integration. In the near future, development on outer loop vectorization will continue, and merging 
-it with LV's VPlan path is planned.
+it with LV's VPlan path is planned. Otherwise, development is currently underway to enable 
+tail-loop folding by default.
 
 Active development of other approaches, such as SLP and whole-function vectorization based on the VPlan 
-architecture, has yet to be started.
+architecture, has yet to be started. Lastly, the initially planned cost model which enabled a standalone 
+VPlan to calculate its own cost has not been implemented yet but could become a vital system in the future 
+of VPlan.
 
-% TODO: Write about potential future work benchmarking
-%       Loop Code Benchmark collection
-%       Quick note about collaborative development (example stability & miscompile policy)
-%       LLVM Loop Optimization as one of the most active subareas of LLVM development
+In the future, it might be interesting to benchmark and compare different auto-vectorization systems, 
+such as the old LLVM LoopVectorize pass, the current state of LLVM with its new VPlan system, 
+GCC~\cite{autovecoptgcc}, ICC, and region vectorizers~\cite{Moll_2021,10.1145/3303117.3306172,rv}. Interesting metrics to 
+compare would be compile time, performance and size of the produced code, differences in the chosen 
+vectorization approaches, differences in legality checks that lead to loops not being vectorized, 
+correctness, and stability. These metrics can then differ based on the target architecture.
+
+Various suitable benchmark sets exist that have also been used for similar 
+comparisons, such as TSVC 2 or SPEC CPU 2017.
 

### 2025-01-28 paper/paper.tex
paper

@@
 Modern CPUs are often equipped with multiple different vector registers. These registers are nowadays 
-as wide as 512 bits, allowing for the processing of multiple data streams at once (SIMD). By 
+as wide as 512 bits or more, allowing for the processing of multiple data streams at once (SIMD). By 
 batching multiple values together in one register, different ISAs like Intel AVX or ARM SVE allow 
@@
 (VPlan)~\cite{llvmextloopvec,llvmvplan}. The final goal would be to unite LLVM's auto-vectorization
-in a single flexible system capable of optimizing SLPs, inner, and nested loops with complex 
+in a single flexible system capable of optimizing SLPs, and inner and nested loops with complex 
 control-flow.
@@
 After that, the best VPlan, and with it, the best vectorization approach is chosen by the cost model.
-This VPlan is then executed by going over its own vectorized control flow representation and
+This VPlan is then executed by going over its own vectorized intermediate representation and
 materializing it back into the underlying LLVM IR.
 
-LLVM's VPlan is currently being integrated into the existing Loop Vectorizer and is
+LLVM's VPlan system is currently being integrated into the existing Loop Vectorizer and is
 already modeling most inner loop vectorizations and transformations~\cite{llvmvplanupdate}. 
@@
 strategies and constraints in general and describe the original auto-vectorization
-approach in LLVM. Section~\ref{sec:vplan} then introduces the new VPlan approach, focusing on
-its most important individual components, and gives an outlook on future developments. In 
-Section~\ref{sec:relatedwork}, we highlight work related to LLVM's auto-vectorization. Concluding,
-we summarize VPlan, its current state, and its potential future.
+approach in LLVM. Section~\ref{sec:vplan} then introduces the new VPlan system, focusing on
+its core VPlan component. Next, we summarize the current state of VPlan as of LLVM version 19.1.4.
+Section~\ref{sec:vplanfuture} gives an outlook on future developments regarding VPlan. In 
+Section~\ref{sec:relatedwork}, we highlight other approaches to auto-vectorization and 
+comparisons of LLVM's auto-vectorization to other compilers such as GCC. 
+Concluding, we summarize the intentions of VPlan, its current state, and its potential future.
 
@@
 Vectorization of these loops is more straightforward as the only option to vectorize is over 
-the loop's induction variable.
+the one loop's induction variables.
 
@@
 inner loop, or a mix of both, representing a hybrid/multi-dimensional vectorization approach.
-Finding the best legal vectorization can therefor be more difficult.
+Finding the best legal vectorization can, therefore, be more difficult.
 
@@
   \caption{A code snippet of a nested loop where it would be beneficial to vectorize over the 
-  outer loop. The inner loop has a trip count (6) that is too low to vectorize. Additionally,
+  outer loop. The inner loop has a trip count that is too low to vectorize (6). Additionally,
   the memory access pattern of \texttt{y} would be consecutive compared to scattered when
@@
 this call could be replaced with a call to a newly constructed function with the same semantic 
-control-flow but operating on multiple values at once through vector registers.
+control-flow but operating on multiple values at once by utilizing vector registers.
 
@@
 Some of the most common vectorization barriers are data dependencies. As the vectorized version
-of a code executes for Vectorization Factor (VF) $N$ data streams simultaneously, it inherently changes
+of a code executes for Vectorization Factor (VF) data streams simultaneously, it inherently changes
 the order of operations. To guarantee that the semantics of the code do not change, all
-operations on these $N$ data streams must, generally, be independent. The simplest case
-where this is not provided is when the $i$-th loop iteration depends on any of the $(i-N)$-th 
-previous iterations.
+operations on these VF data streams must, generally, be independent. The simplest case
+where this is not provided is when the $i$-th loop iteration depends on any of the 
+previous iterations up to the $(i-\text{VF})$-th.
 
@@
 loop is executed, can often not be vectorized. This is because their exit condition usually depends on 
-the loop's body itself, making it impossible to predetermine how many vector lanes are needed for the next 
-loop iteration when executing.
+the loop's body itself, making it impossible to predetermine how many iterations could be processed together 
+before having to check the exit condition again.
 
@@
 instruction equivalent. For example Integer division, commonly implemented using the 
-\textit{Newton-Raphson method}, is already a complex and iterative operation 
+Newton-Raphson method, is already a complex and iterative operation 
 on modern CPUs. Applying this algorithm 
@@
 \subsection{Vectorization Costs}
-While vectorizing a given control flow may not be legal, the vectorized code sometimes performs 
+While vectorizing a given control flow may not be legal, the vectorized code can sometimes also perform 
 worse than its scalar equivalent.
@@
 would then be able to guess the correct branch in the scalar code. The vectorized version would
-always load the \texttt{y[i]} into memory only for the addition to be masked away. Different 
+always access the memory at \texttt{y[i]} only for the addition to be masked away. Different 
 approaches to avoid such runtime penalties are branch-on-superword-condition code 
@@
 For example, gather, scatter, 
-shuffle, and permutation instructions are typically slow. Control flow with many 
+shuffle, and permutation instructions are typically considered slow. Control flow with many 
 non-adjacent memory accesses or 
 operations that combine different vector elements from the same register (horizontal reductions) can thus 
-decrease vectorization performance.
+decrease overall performance.
 
@@
 due to many factors: (1) The transformed control flow often includes more instructions.
-(2) As the loop trip count might not be a perfect multiple of the Vectorization Factor, a scalar
+(2) As the loop trip count might not be a perfect multiple of the vectorization factor, a scalar
 epilogue/tail of the loop might be used to handle the remaining loop iterations. (3) Various runtime
 checks might be inserted throughout the vectorized control flow to handle, for example, pointers 
-to overlapping memory regions, in which case the scalar version of the loop must be executed due to 
-unclear data dependencies.
+to overlapping memory regions, in which case the scalar version of the loop must be executed.
 
@@
 and nested loop vectorizations would have been difficult, as LV was designed to flatten a
-loop's control flow into a single basic block. Producing and considering more and 
-more artifacts when trying to transform the scalar loop into a vectorized one all at once also did 
-not scale well.
+loop's control flow into a single basic block. Producing and considering more artifacts when trying to 
+transform the scalar loop into a vectorized all at once also did not scale well.
 
 As a potential solution, Intel proposed a long-term effort to refactor LLVM's whole auto-vectorization
-using a more modular and flexible system called Vectorization Plan (VPlan)~\cite{llvmextloopvec}.
+pipeline to use a more modular and scalable system called Vectorization Plan (VPlan)~\cite{llvmextloopvec}.
 
@@
 Intel's initial VPlan proposal aimed to unite any 
-auto-vectorization systems (SLP, Inner/Outer-Loop, Whole-Function) into a 
+auto-vectorization strategy (SLP, Inner/Outer-Loop, Whole-Function) into a 
 single flexible system~\cite{llvmextloopvec,llvmintrvplan,llvmvplanstate}. 
-This system would use VPlans to represent different 
+This system would use VPlan objects to represent different 
 vectorization approaches, abstracting away from the underlying scalar loop IR using its own 
 hierarchical control flow graph (H-CFG). This allows iteratively exploring different 
-optimization decisions by performing transformations on this CFG without modifying any LLVM IR.
-Each VPlan could to calculate its cost and execute itself, materializing the vectorization 
-approach back into LLVM IR.
+optimization decisions by performing transformations on this graph without modifying the 
+underlying LLVM IR.
+Each VPlan could calculate its cost and execute itself, materializing its vectorization 
+approach back into the LLVM IR.
 
@@
 Before explaining how far this system has been implemented in LLVM, we will review
-the different elements of a \texttt{VPlan} in more detail.
+the different elements of a \textit{VPlan} in more detail.
 
@@
 \paragraph{H-CFG}
-\emergencystretch 1em
+\emergencystretch 2em
 
 This is achieved by introducing a new layer of abstraction from the underlying loop's scalar IR.
-Each VPlan stores its own Hierarchical-CFG (H-CFG) modeling the vectorization candidate. This 
+Each VPlan stores its own Hierarchical-CFG (H-CFG) modeling its vectorization candidate. This 
 control-flow graph and its components are the main elements of a single VPlan. It is hierarchical
@@
   the LLVM IR's \texttt{BasicBlock} holding a list of IR Instructions.
-  \item \texttt{VPRegionBlock} represents a subgraph with a single VP entry node and a 
-  single VP exit node. A \texttt{VPRegionBlock} can be marked to replicate its subgraph
-  \texttt{VF * UF} times when executing. This is useful for modeling scalarized or predicated
+  \item \texttt{VPRegionBlock} represents a subgraph with a single entry and exit node. 
+  A \texttt{VPRegionBlock} can be marked to replicate its subgraph
+  \texttt{VF~*~UF} times when executing. This is useful for modeling scalarized or predicated
   instructions within the final vectorized loop body.
@@
   \includegraphics[width=0.475\textwidth]{images/inner-loop-vplan-hcfg-loop-body-color.png}
-  \caption{The vectorized loop part of the hierarchical CFG of a VPlan modeling the 
+  \caption{The vectorized loop part of a VPlan's hierarchical CFG modeling the 
   conditional loop depicted in Figure~\ref{fig:inner-loop-vec}. 
-  Surrounding \texttt{VPIRBasicBlock}-nodes are omitted.
+  Surrounding nodes were omitted for clarity.
   Some predefined values are
-  \texttt{vp<\%0>} = VF * UF; \texttt{vp<\%1>} = vector trip count; \texttt{vp<\%2>} = original trip count.}
+  \texttt{vp<\%0>} = \texttt{VF * UF}; \texttt{vp<\%1>} = vector trip count; \texttt{vp<\%2>} = original trip count.}
   \label{fig:inner-loop-vplan-hcfg-body}
@@
 of Figure~\ref{fig:inner-loop-vec}. Comparatively, Figure~\ref{fig:inner-loop-vplan-hcfg-body}
-shows the vectorized loop part of the H-CFG of a VPlan that models the vectorization approach described
-in Figure~\ref{fig:inner-loop-vec}. Note that the full H-CFG of this VPlan consists of many more nodes 
+shows the equivalent vector loop part of a VPlan's H-CFG that models the vectorization approach described
+in Figure~\ref{fig:inner-loop-vec}. Note that the complete H-CFG of this VPlan consists of many more nodes 
 to handle legality checks (coming before the vectorized loop) or the entire scalar epilogue loop 
@@
 \texttt{vector.preheader} and \texttt{middle.block} 
-are simple \texttt{VPBasicBlocks} surrounding the main vector loop node. This \texttt{vector loop} node 
+are simple \texttt{VPBasicBlocks} surrounding the main \texttt{vector loop} node. The \texttt{vector loop} node 
 is a \texttt{VPRegionBlock} that does 
@@
 
-\texttt{VPRecipes} can be seen as instructions of an IR specialized for auto-vectorization. 
+\texttt{VPRecipes} can be seen as instructions of an intermediate representation specialized for auto-vectorization. 
 A single \texttt{VPRecipe} inside a VPlan is based on zero or more LLVM IR instructions 
-of the underlying scalar IR, called ingredients, and will materialize into one or more LLVM IR 
-instructions when executed.
+of the underlying scalar IR. These underlying IR instructions are called \textit{ingredients}. 
+Each \texttt{VPRecipe} will materialize back into one or more LLVM IR instructions when it is executed.
 
@@
   \item \texttt{VPReplicateRecipe} will replicate one IR instruction either once (\texttt{CLONE})
-  or VF * UF times (\texttt{REPLICATE}) into the vectorized IR when executed.
+  or \texttt{VF * UF} times (\texttt{REPLICATE}) into the vectorized IR when executed.
   \item \texttt{VPWidenRecipe} will \texttt{WIDEN} one IR instruction into a
@@
 Most of the \texttt{VPRecipes} in Figure~\ref{fig:inner-loop-vplan-hcfg-body} have capitalized 
-\texttt{HINTS} indicating their type. Additionally instructions/recipes of the same color 
-between Figure~\ref{fig:inner-loop-scalar-ir} and Figure~\ref{fig:inner-loop-vplan-hcfg-body}
+\texttt{HINTS} indicating their type. Additionally IR instructions/\texttt{VPRecipes} of the same color 
+in Figure~\ref{fig:inner-loop-scalar-ir} and Figure~\ref{fig:inner-loop-vplan-hcfg-body}
 are related in their semantical effect.
 
-The only recipes in our example VPlan CFG that do not semantically model any 
+The only \texttt{VPRecipes} in our example VPlan H-CFG that do not semantically model any 
 instructions found in the scalar IR are from the \texttt{middle.block} (purple). 
-This block decides if the scalar epilogue loop is executed after the vectorized loop by 
+This block decides if the scalar epilogue loop is still executed after the vectorized loop by 
 comparing the vector trip count to the original loop trip count. This additional vector trip count is 
-needed as the original loop trip count might not be a perfect multiple of VF * UF, the number of elements 
-processed in a single iteration.
+needed as the original loop trip count might not be a perfect multiple of \texttt{VF * UF}, the number of elements 
+processed in a single vector loop iteration.
 
 The red recipes generally handle the induction variable \texttt{i} and the termination of the 
-vectorized loop. \texttt{vp<\%3>} is this loop's scalar induction variable, which goes from 0 to 
+vector loop. \texttt{vp<\%3>} is this loop's scalar induction variable, which goes from 0 to 
 \texttt{vp<\%1>}, the vector trip count. The per vector-lane offset's step size is 1 (\texttt{ir<1>}), 
 meaning that each 
-vector-lane offset from the base induction variable is one greater than the previous lane (we access the 
-memory consecutively).
+vector-lane offset from the base induction variable (\texttt{vp<\%3>}) is one greater than the previous lane. 
+Thus the memory is accessed consecutively.
 
-The blue recipes model the condition guarding the addition. \texttt{vp<\%6>} holds the mask, which 
-results from this condition and is used for the guarded memory operations.
+The blue recipes model the condition guarding the addition. \texttt{vp<\%6>} holds the conditional mask 
+and is used for the guarded memory operations.
 
 Finally, the green recipes represent the addition. First, the data from both \texttt{y} and \texttt{x} 
-gets loaded (guarded by the compare mask), then added together, and then stored back into 
-\texttt{x} (again, guarded by the compare mask).
+gets loaded, then added together, and then stored back into \texttt{x}.
 
@@
   redundant type casts, and simplifying logical patterns.
-  \item Removing dead \texttt{VPRecipes} whose values have no users.
+  \item Removing dead \texttt{VPRecipes} whose produced values have no users consuming them.
   \item Moving loop-invariant \texttt{VPRecipes} out of the vector loop region. A \texttt{VPRecipe} is 
@@
 In addition to the general vectorization legality concepts mentioned in 
-Section~\ref{sec:vec-legal}, LV has a several extra conditions to check.
+Section~\ref{sec:vec-legal}, LoopVectorize has a several extra conditions to check.
 
@@
   An indirect branch is a branch that jumps to a value evaluated at runtime. A back-edge in a loop 
-  is an edge from the loop back to it. 
+  is an edge from the loop's body back to its start. 
   \item Loops with non-linear control flow must be if-convertible. This means any branch or switch 
-  statement must not exit the loop. All instructions of such a block can be executed speculatively 
+  statement must not exit the loop. All instructions of such a predicated block can be executed speculatively 
   (e.g. via masking)
@@
 Additionally, during this phase, all recurrences of the loop, such as reductions or inductions, are analyzed,
-and their phi nodes are stored in a set.
+and their phi nodes are stored for the later initialization of VPlans.
 
 \subsection{Phase 2: Planning}
-After the Legality Phase has analyzed the scalar loop and it is deemed legal to vectorize the 
-\texttt{LoopVectorizationCostModel} and the \texttt{LoopVectorizationPlanner} are initialized.
-The latter one then constructs the initial VPlans based on the properties gathered during the legal phase
-and the cost model initialization, such as the recurrence variables or the maximum viable vectorization 
-factor.
+After the Legality Phase has analyzed the scalar loop and it was deemed legal to vectorize, the 
+\texttt{LoopVectorizationCostModel} and \texttt{LoopVectorizationPlanner} are initialized.
+The latter one then constructs the initial VPlans for different vectorization factors based on the properties 
+gathered during the legal phase
+and the cost model initialization. These properties are, for example, recurrence variables or the maximum viable 
+vectorization factor.
 
@@
 region block for the vector loop, a middle block with the potential runtime check for the scalar 
-epilogue loop, and a block for the scalar loop. After the H-CFG has been created, the scalar LLVM IR 
+epilogue loop, and a block for the scalar epilogue loop itself. Then the scalar LLVM IR 
 is iterated in topological order, and each scalar IR instruction is transformed into a \texttt{VPRecipe}.
-After modeling each scalar IR instruction with a \texttt{VPRecipe}, several VPlan transformations 
-are executed, optimizing the initialized VPlan.
+After that the original scalar loop is represented by a general vectorization approach. Lastly, several VPlan 
+transformations are executed, transforming and optimizing the VPlan's H-CFG and its \texttt{VPRecipes}.
 
-After the initial VPlans have been optimized, the best Vectorization Factor is computed. This is done 
+After all the initial VPlans have been optimized, the best vectorization factor is computed. This is done 
 by comparing all generated VPlans for all their possible VFs and choosing the most profitable one. If 
@@
 Currently, LV and VPlan still mostly rely on the old cost model. This legacy cost model derives costs 
-based on the scalar LLVM IR instructions or, if applicable, their vectorized equivalent.
+based on the scalar LLVM IR instructions or, if applicable, their direct vectorized equivalent.
 This is suboptimal for VPlan as VPlans work on \texttt{VPRecipes} abstracted away from the underlying 
@@
 A remainder of the old \texttt{LoopVectorize} implementation before VPlan is the 
-before-mentioned \texttt{LoopVectorizationCostModel}. It is a single object that computes different costs based on the scalar loop 
+before-mentioned \texttt{LoopVectorizationCostModel}. It is a single object that computes costs based on the scalar loop 
 IR without any relation to different VPlans. For example, it stores the instructions that will remain scalar 
@@
 On the other hand, \texttt{VPRecipes} that do not directly model an underlying LLVM scalar IR instruction, 
-such as \texttt{VPInstruction}, can generally not compute their own cost and thus return a cost of zero 
+such as \texttt{VPInstruction}, can generally not compute their own cost yet, thus returning a cost of zero 
 for now.
@@
 outer/hybrid loop vectorization are implemented, this new cost model would be an important
-detail of the VPlan system to reliably choose the most optimal vectorization approach.
+detail of the VPlan system to reliably choose the most optimal vectorization candidate.
 
 \subsection{Phase 3: Execution}
-After the best VPlan has been chosen, it is executed (\texttt{LoopVectorizationPlanner::executePlan}).
+After the best VPlan has been chosen, it is executed.
 For this VPlan, some final transformations are run depending on a specifically chosen VF and UF.
 Then a vectorized code skeleton is created in the IR, and the VPlan executed. 
-Upon execution, the H-CFG of the VPlan is walked, and each node is executed, which in turn executes all 
+The H-CFG of the VPlan is walked, and each node executed, which in turn executes all 
 of its \texttt{VPRecipes} in order. Each \texttt{VPRecipe} produces its respective output IR.
@@
 A \texttt{VPTransformState} object is passed along to store information 
-needed for generating the output IR, such as the chosen Vectorization Factor or generated values that were 
+needed for generating the output IR, such as the chosen vectorization factor or generated values that were 
 used by later \texttt{VPRecipes}.
@@
 While Figure~\ref{fig:inner-loop-vec-cfg} shows the general CFG of a vectorized loop in LLVM, some 
-nodes can also be omitted in special cases.
+of its nodes can also be omitted in special cases.
 
@@
 is guaranteed that they will not overlap and the \texttt{vector.memcheck} block will be omitted.
-While the C language and most modern C++ compilers have an explicit \texttt{restrict} keyword for this,
+While the C language and most modern C\texttt{++} compilers have an explicit \texttt{restrict} keyword for this,
 Rust automatically applies the \texttt{noalias} attribute to safe references when compiling via LLVM.
@@
 Additionally, before the initial VPlans are constructed at the beginning of Phase 2, it is also checked if 
-the scalar epilogue loop could be folded into the vectorized loop version itself by increasing the vector 
+the scalar epilogue loop could be folded into the vector loop itself by increasing the vector 
 trip count over the actual loop trip count and masking all operations. This tail-loop folding approach in LLVM 
@@
 
-Otherwise, if the scalar epilogue loop could be vectorized, it would be in a second loop-vectorize 
+Otherwise, if the newly created scalar epilogue loop could be vectorized, it would be in a second loop-vectorize 
 pass by, again, creating VPlans, finding the best one, and executing it.
@@
 \section{Future of VPlan}
+\label{sec:vplanfuture}
+
 Many of the VPlan internals and surrounding systems are still in active development. To stay up-to-date 
 with the latest developments regarding auto-vectorization in LLVM, one can follow the 
-Community Forum\footnote{\url{https://discourse.llvm.org/c/ir-optimizations/loop-optimizations/62}} or
+community forum\footnote{\url{https://discourse.llvm.org/c/ir-optimizations/loop-optimizations/62}} or
 attend the monthly online 
@@
 explicit VPlan-to-VPlan transformations. Additionally, the legacy LV cost model still influences 
-many cost model-related decisions. However, the transition to a VPlan-based cost model has already 
+many cost-related decisions. However, the transition to a VPlan-based cost model has been 
 started. This new model would derive costs based on \texttt{VPRecipes} directly instead of the 
@@
 
-Feng et al. concluded that auto-vectorizers were mostly limited by analysis capabilities 
+Feng et al. concluded that auto-vectorizers were mostly limited by analyzing capabilities 
 needed to reason about specific optimization approaches and inaccurate cost models.
@@
 similar between GCC 14 and LLVM 17, with GCC only being marginally better. LLVM 17 tends to favor 
-generating vector code targeting specific vector sizes. However, it is unclear whether this is an 
+generating vector code targeting specific vector sizes. However, it is unclear whether this has an 
 actual drawback.
@@
 By now, VPlan has been in active development for several years. 
-This system would use different VPlan objects to efficiently 
+This system would use multiple VPlan objects to efficiently 
 model and explore different vectorization approaches simultaneously.
-It adds a new layer using \texttt{VPRecipes} to model 
+It adds another layer of abstraction by using \texttt{VPRecipes} to model 
 vectorization-specific instructions over the underlying LLVM IR. 
-This abstraction allows for efficiently exploring different optimization decisions simultaneously, 
-keeping the underlying implementation flexible and modular. In the end, a cost model currently still 
+This allows for efficiently exploring multiple optimization decisions simultaneously, 
+keeping the underlying implementation flexible and modular. In the end, a cost model currently 
 working based on LLVM IR instead of \texttt{VPRecipes} is used to compare different VPlans 
@@
 
-Outer loop vectorization support via VPlan is still experimental and separated from the inner loop
-LV integration. In the near future, development on outer loop vectorization will continue, and merging 
-it with LV's VPlan path is planned. Otherwise, development is currently underway to enable 
+While inner loop vectorization is already implemented, support for outer loops is still experimental. 
+In the near future, development on outer loop vectorization will continue, and merging 
+it with the LoopVectorize VPlan path is planned. Besides that, development is currently focusing on enabling 
 tail-loop folding by default.
@@
 
-In the future, it might be interesting to benchmark and compare different auto-vectorization systems, 
-such as the old LLVM LoopVectorize pass, the current state of LLVM with its new VPlan system, 
-GCC~\cite{autovecoptgcc}, ICC, and region vectorizers~\cite{Moll_2021,10.1145/3303117.3306172,rv}. Interesting metrics to 
+To help evaluate the idea and current state of VPlan, it might be interesting to benchmark and 
+compare it against different auto-vectorization systems, 
+such as LLVM before VPlan, 
+GCC, ICC, and region vectorizers~\cite{Moll_2021,10.1145/3303117.3306172,rv}. Interesting 
+metrics to 
 compare would be compile time, performance and size of the produced code, differences in the chosen 
@@
 comparisons, such as TSVC 2 or SPEC CPU 2017.
+
+In conclusion, the new VPlan architecture has already transformed LLVM's previously rigid auto-vectorization 
+pipeline into a much more modular and flexible system. While some important features are still missing, the 
+LLVM community has continuously improved VPlan over the last few years and will continue to do so.
 
